---
title: "Example FUSION2 Lidar Pipeline: Ground Filter, Canopy Model, Grid Metrics"
author: "Jacob Strunk"
date: today
format: html
execute:
  echo: true
  warning: false
  message: false
---

<!--
Pipeline script -- runs an ordered sequence of FUSION2 command-line tools
against a small example lidar point cloud tile. Not a scientific finding, so
this follows the pipeline template (execution-ordered "# Step N" headings),
not the manuscript-style analysis template.
-->

# Setup

## Reset working directory to the project root

<!--
This script lives in R/analysis/, but Quarto executes chunks with the
working directory set to the .qmd's own folder by default -- so
project-relative paths like "data/raw" would otherwise resolve against
R/analysis/, not the project root. root.dir only takes effect starting
with the chunk after it is set.
-->

```{r}
#| label: root-dir

knitr::opts_knit$set(root.dir = "../..")
```

## Load libraries and set default options

```{r}
#| label: setup

# data.table for the small metrics summary table; terra for reading back
# and previewing the GeoTIFF rasters the FUSION2 tools write; RForInvt for
# output versioning; ggplot2 for the one preview figure; DBI/RSQLite for
# reading gridmetrics' SQLite table-output option (Step 4) back into R.
library(data.table)
library(terra)
library(RForInvt)
library(ggplot2)
library(DBI)
library(RSQLite)

knitr::opts_chunk$set(
    echo      = TRUE
  , warning   = FALSE
  , message   = FALSE
  )

options(scipen = 10E6)

# Directories used throughout this pipeline.
dir_lidar_raw = file.path("data", "raw", "lidar")
dir_derived   = file.path("data", "derived", "lidar_example")
dir_tables    = file.path("output", "tables")
dir.create(dir_derived, recursive = TRUE, showWarnings = FALSE)
dir.create(dir_tables, recursive = TRUE, showWarnings = FALSE)

# The compiled FUSION2 CLI tools are not bundled with this repo -- build
# FUSION2 from source (https://github.com/jstrunk001/FUSION2) and either put
# its bin/ folder on PATH or point the FUSION2_BIN_DIR environment variable
# at it.
fn_find_fusion_tool = function(tool_name) {
  path_on_path = Sys.which(tool_name)
  if (nzchar(path_on_path)) return(normalizePath(path_on_path, winslash = "/"))
  dir_env = Sys.getenv("FUSION2_BIN_DIR")
  if (nzchar(dir_env)) {
    path_in_env_dir = file.path(dir_env, tool_name)
    if (file.exists(path_in_env_dir)) return(normalizePath(path_in_env_dir, winslash = "/"))
  }
  stop(
      "Could not find ", tool_name, ". Build FUSION2 from source "
    , "(https://github.com/jstrunk001/FUSION2) and either add its bin/ "
    , "folder to PATH or set the FUSION2_BIN_DIR environment variable to it."
    )
}
path_groundfilter = fn_find_fusion_tool("groundfilter.exe")
path_canopymodel  = fn_find_fusion_tool("canopymodel.exe")
path_gridmetrics  = fn_find_fusion_tool("gridmetrics.exe")
path_cloudmetrics = fn_find_fusion_tool("cloudmetrics.exe")

# The small example point cloud tile this pipeline runs against. See the
# "Pipeline overview" section below for how it was sourced and reduced.
path_las_example = file.path(
    dir_lidar_raw
  , "TC_1372_forest200ft.laz"
  )
if(!file.exists(path_las_example)) stop("Example lidar tile not found at ", path_las_example)
```

## Define helper functions used throughout this pipeline

```{r}
#| label: helper-functions

#1. define a helper that runs one FUSION2 CLI tool and reports what happened
#  - wraps system2() in try() per r_coding_standards.md Section 16 (no
#    tryCatch())
#  - returns the captured console output invisibly, and stops with a clear
#    message if the tool exits with a non-zero status
fn_run_fusion_tool = function(exe_path, args, tool_label){
  if(!file.exists(exe_path)) stop("FUSION2 tool not found: ", exe_path)
  result = try(
      system2(exe_path, args = args, stdout = TRUE, stderr = TRUE)
    , silent = TRUE
    )
  exit_status = attr(result, "status")
  has_error   = inherits(result, "try-error") || (!is.null(exit_status) && exit_status != 0)
  cat(paste(result, collapse = "\n"), "\n")
  if(has_error) stop(tool_label, " exited with an error -- see console output above.")
  invisible(result)
  }

#2. define a helper that summarizes one GeoTIFF raster's cell values into a
#   single-row data.table -- reused across the three rasters this pipeline
#   produces
fn_summarize_raster = function(path_raster, layer_name){
  r = terra::rast(path_raster)
  v = terra::values(r[[layer_name]])
  data.table::data.table(
      layer   = layer_name
    , n_cells = length(v)
    , n_na    = sum(is.na(v))
    , min_val = min(v, na.rm = TRUE)
    , mean_val = mean(v, na.rm = TRUE)
    , max_val = max(v, na.rm = TRUE)
    )
  }
```

# Pipeline overview

This pipeline demonstrates a standard ground-classification-to-metrics
lidar workflow using four FUSION2 command-line tools -- `groundfilter`,
`canopymodel`, `gridmetrics`, and `cloudmetrics` -- against one small, real
example point cloud tile. The `gridmetrics` and `cloudmetrics` steps also
exercise the current full metric bundle: height/intensity strata
(`/strata`/`/intstrata`), surface roughness (`/surfstats`), and
`gridmetrics`' opt-in per-cell table export (`/output-table`) in both of
its supported formats, CSV and SQLite.

**Example tile source.** The input tile is a 200 x 200 ft window of
closed-canopy forest clipped from lidar tile `TC_1372` (NV5 Geospatial,
2022 acquisition; 2250 x 2250 ft, about 6.2 million points). The window
has 70,532 points, 92% canopy cover, a 95th-percentile height of 142 ft,
and a tallest return of 176 ft. It is kept in the tile's native LAS 1.4
point format 6 LAZ encoding (about 0.5 MB). The projected coordinate
system is NAD83(HARN) / Washington South (ftUS), EPSG:2927, vertical datum
NAVD88 (ftUS, Geoid03). `006_prepare_forested_example_tile.qmd` shows how
the window was chosen: it scores every 200 ft window of the full tile by
canopy cover and height and keeps the tallest window with at least 70%
cover.

Steps:

1. Classify ground points and build a ground surface DEM (`groundfilter`).
2. Build a canopy height model from the point cloud and ground DEM
   (`canopymodel`).
3. Compute a grid of the full lidar elevation/intensity metric bundle,
   height/intensity strata, and surface roughness (`gridmetrics`),
   exporting a per-cell CSV table alongside the raster.
4. Re-run `gridmetrics` as a table-only export (`/noraster` +
   `/output-table:<path.sqlite>`), demonstrating the SQLite table format.
5. Summarize the derived rasters and tables into one small metrics table
   and preview the canopy height model.
6. Compute per-plot summary and height-strata metrics with `cloudmetrics`.

# Step 1: Classify ground points and build a ground surface DEM

`groundfilter` separates ground returns from the rest of the point cloud
and interpolates them into a ground surface raster, which the next two
steps use to normalize point and canopy heights above ground.

```{r}
#| label: step-1-groundfilter

#1. mint a versioned output path for the ground DEM
path_ground_dem = RForInvt::file_version(
    file.path(dir_derived, "ground_dem.tif")
  , increment = TRUE
  , note = "ground DEM from the forested TC_1372 200x200 ft clip"
  )

#2. run groundfilter against the example tile
#  - the CLI-harmonization pass renamed this flag from /output-dem: to
#    /output-raster:; the old name is still silently accepted (exit 0)
#    but produces no output file at all, so this was previously a silent
#    no-op -- confirmed by direct testing against the current build
fn_run_fusion_tool(
    path_groundfilter
  , args = c(
        path_las_example
      , paste0("/output-raster:", path_ground_dem)
      , "/cellsize:5"
      )
  , tool_label = "groundfilter"
  )
```

# Step 2: Build a canopy height model

`canopymodel` builds a top-of-canopy surface from the point cloud and
normalizes it to height above ground using the DEM from Step 1.

```{r}
#| label: step-2-canopymodel

#1. mint a versioned output path for the canopy height model
path_chm = RForInvt::file_version(
    file.path(dir_derived, "chm.tif")
  , increment = TRUE
  , note = "canopy height model from the forested TC_1372 200x200 ft clip"
  )

#2. run canopymodel against the example tile, using Step 1's ground DEM
fn_run_fusion_tool(
    path_canopymodel
  , args = c(
        path_las_example
      , paste0("/output:", path_chm)
      , paste0("/ground:", path_ground_dem)
      , "/cellsize:5"
      )
  , tool_label = "canopymodel"
  )
```

# Step 3: Compute a grid of the full metric bundle, strata, and surface roughness

`gridmetrics` computes a per-cell raster of the full lidar elevation/
intensity statistic bundle over the example tile -- min/max/mean/stddev/
variance/skewness/kurtosis, the full percentile curve, L-moments, and more
(see `METRICS_REFERENCE.md` Section 1 for the complete column list), again
normalized using Step 1's ground DEM. Three further capabilities are
requested alongside that base bundle:

- `/strata:0.5,2.0,5.0,10.0,20.0,50.0,100.0` buckets each cell's points into eight
  height strata (the seven thresholds given, plus one "above the last
  threshold" bucket) and reports a count/proportion/mean/stddev/min/max
  summary per bucket, both as CSV columns and, with `/strataraster`, as
  raster bands (`stratum_00_mean`, `density_stratum_00`, ...).
  `/intstrata` (omitted here, so it defaults to `/strata`'s thresholds)
  would report the same buckets' **intensity** values instead of
  elevation.
- `/surfstats` adds `surface_area_ratio` and `roughness` bands, derived
  directly from the per-cell elevation grid `gridmetrics` already holds in
  memory -- no second raster read. `/surfstats-source:max` (the default,
  named explicitly here) feeds it each cell's `elev_max`, appropriate
  since this grid holds ground-normalized canopy heights, not raw ground
  elevations.
- `/output-table:<path.csv>` writes that same full per-cell table as a CSV
  file alongside the raster -- opt-in, so nothing is written unless the
  option is given. Step 4 below re-runs `gridmetrics` with `/noraster` and
  a `.sqlite` path instead, to demonstrate the table-only, SQLite side of
  the same option.

`/nodata:` and `/noheight:` are named explicitly below even though both are
left at their defaults, so a reader sees them: `/nodata:NA` marks a grid
cell with zero lidar returns at all (a genuine data gap -- nothing landed
there), while `/noheight:0` marks a cell (or stratum bucket) that has
returns but none clear the `/minht` cutoff (a real "no canopy here" answer,
not missing data). `terra::rast()` already treats a GeoTIFF's registered
NoData value (including `NaN`, which `/nodata:NA` resolves to) as `NA` on
read, and `data.table::fread()`'s default `na.strings` already includes
`"NA"` for the CSV table -- both example-script readers below pick up the
convention automatically, no reader-side code change needed.

```{r}
#| label: step-3-gridmetrics

#1. gridmetrics writes its own output filename into outdir, based on the
#   input file's name -- point it at a dedicated subfolder so the result
#   is easy to find and version
dir_gridmetrics_out = file.path(dir_derived, "gridmetrics")
dir.create(dir_gridmetrics_out, recursive = TRUE, showWarnings = FALSE)

#2. mint a versioned path for the per-cell CSV table /output-table writes
#   alongside the raster
path_gridmetrics_table = RForInvt::file_version(
    file.path(dir_tables, "gridmetrics_table.csv")
  , increment = TRUE
  , note = "gridmetrics per-cell CSV table (full metric bundle plus strata) from the forested TC_1372 200x200 ft clip"
  )

#3. run gridmetrics against the example tile, requesting the full metric
#   bundle's height strata, per-stratum raster bands, surface roughness,
#   and a per-cell CSV table alongside the raster
fn_run_fusion_tool(
    path_gridmetrics
  , args = c(
        path_las_example
      , paste0("/ground:", path_ground_dem)
      , "/cellsize:10"
      , paste0("/outdir:", dir_gridmetrics_out)
      , "/nodata:NA"
      , "/noheight:0"
      , "/strata:0.5,2.0,5.0,10.0,20.0,50.0,100.0"
      , "/strataraster"
      , "/surfstats"
      , "/surfstats-source:max"
      , paste0("/output-table:", path_gridmetrics_table)
      )
  , tool_label = "gridmetrics"
  )

#4. mint a versioned copy of the resulting raster at a stable name
path_gridmetrics_raw = list.files(
    dir_gridmetrics_out
  , pattern = "_gridmetrics\\.tif$"
  , full.names = TRUE
  )[1]
path_gridmetrics_raster = RForInvt::file_version(
    file.path(dir_derived, "gridmetrics.tif")
  , increment = TRUE
  , note = "gridmetrics raster (full metric bundle, strata, surface stats) from the forested TC_1372 200x200 ft clip"
  )
file.copy(path_gridmetrics_raw, path_gridmetrics_raster, overwrite = TRUE)
```

# Step 4: Re-run `gridmetrics` as a table-only export (SQLite, no raster)

`/noraster` skips the GeoTIFF entirely -- FUSION2 requires `/output-table`
alongside it, since a run must produce at least one output. This re-run of
Step 3 writes the identical full metric bundle and strata summary as a
`.sqlite` database instead of a raster, demonstrating the second of the
two table formats `/output-table` accepts (`TableWriter` detects the format
from the path's extension -- `.csv`, or `.sqlite`/`.db`/`.sqlite3`).

```{r}
#| label: step-4-gridmetrics-table-only

#1. mint a versioned path for the SQLite table
path_gridmetrics_sqlite = RForInvt::file_version(
    file.path(dir_tables, "gridmetrics_table.sqlite")
  , increment = TRUE
  , note = "gridmetrics per-cell SQLite table (table-only run, no raster) from the forested TC_1372 200x200 ft clip"
  )

#2. run gridmetrics again with /noraster + /output-table:<path.sqlite> --
#   same inputs and strata options as Step 3, just no GeoTIFF this time
dir_gridmetrics_table_only_out = file.path(dir_derived, "gridmetrics_table_only")
dir.create(dir_gridmetrics_table_only_out, recursive = TRUE, showWarnings = FALSE)

fn_run_fusion_tool(
    path_gridmetrics
  , args = c(
        path_las_example
      , paste0("/ground:", path_ground_dem)
      , "/cellsize:10"
      , paste0("/outdir:", dir_gridmetrics_table_only_out)
      , "/strata:0.5,2.0,5.0,10.0,20.0,50.0,100.0"
      , "/noraster"
      , paste0("/output-table:", path_gridmetrics_sqlite)
      )
  , tool_label = "gridmetrics (table-only, SQLite)"
  )

#3. confirm no raster was written alongside the table -- /noraster's whole
#   point is to skip it
cat(
    "GeoTIFF written in table-only outdir? "
  , length(list.files(dir_gridmetrics_table_only_out, pattern = "\\.tif$")) > 0
  , "\n"
  )

#4. read back the SQLite table via DBI/RSQLite and preview a few columns,
#   including a stratum bucket, to confirm the second table format carries
#   the same content the CSV table (Step 3) and raster bands do
con_gridmetrics_sqlite = DBI::dbConnect(RSQLite::SQLite(), path_gridmetrics_sqlite)
dt_gridmetrics_sqlite = data.table::setDT(DBI::dbReadTable(con_gridmetrics_sqlite, "grid"))
DBI::dbDisconnect(con_gridmetrics_sqlite)
print(dt_gridmetrics_sqlite[ , .(Col, Row, elev_mean, CanopyCover, stratum_02_mean, stratum_02_count)])
```

# Step 5: Summarize the derived rasters and tables, and preview the canopy height model

This step confirms the derived rasters carry real, non-empty values (not
just placeholder output) -- including the new strata bands from Step 3 --
cross-checks Step 3's CSV table against the raster, then saves a small
summary table and a quick preview figure of the canopy height model.

`surface_area_ratio` and `roughness` are computed only from cells with at
least one point above `/minht` (see Step 3) and a full 3 x 3 neighbourhood
of such cells. Under this clip's closed canopy (`canopy_cover` mean about
86%, per the `dt_summary` rows below) every cell qualifies, so both bands
carry values in all 400 cells; on sparse canopy, cells without canopy
neighbours are written as `/nodata`.

```{r}
#| label: step-5-summarize

#1. summarize each raster's cell values into one small comparison table
dt_summary = data.table::rbindlist(list(
    fn_summarize_raster(path_ground_dem, "ground_elevation")
  , fn_summarize_raster(path_chm, "canopy_height")
  , fn_summarize_raster(path_gridmetrics_raster, "elev_mean")
  , fn_summarize_raster(path_gridmetrics_raster, "canopy_cover")
  , fn_summarize_raster(path_gridmetrics_raster, "point_density")
  # elev_aad (average absolute deviation) and r1count (first-return count)
  # are part of gridmetrics' full elevation-statistic bundle and
  # per-return-number counts -- summarized here to confirm those bands
  # exist and carry real values, not just the handful of bands already
  # checked above
  , fn_summarize_raster(path_gridmetrics_raster, "elev_aad")
  , fn_summarize_raster(path_gridmetrics_raster, "r1count")
  # /strataraster and /surfstats bands from Step 3 -- confirms the
  # height-strata buckets and surface-roughness metrics landed in the
  # raster, not just in Step 3's CSV table
  , fn_summarize_raster(path_gridmetrics_raster, "stratum_02_mean")
  , fn_summarize_raster(path_gridmetrics_raster, "density_stratum_02")
  , fn_summarize_raster(path_gridmetrics_raster, "surface_area_ratio")
  , fn_summarize_raster(path_gridmetrics_raster, "roughness")
  ))
print(dt_summary)

#2. save the summary table as a small, tracked output
path_summary_table = RForInvt::file_version(
    file.path(dir_tables, "lidar_example_raster_summary.csv")
  , increment = TRUE
  , note = "raster summary stats for the example lidar pipeline"
  )
data.table::fwrite(dt_summary, path_summary_table)

#3. spot-check Step 3's CSV table against the raster bands summarized
#   above -- the same stratum_02_mean values should show up as both a
#   raster band and a CSV column, since /strata, /strataraster, and
#   /output-table all draw from the same per-cell computation
dt_gridmetrics_table = data.table::fread(path_gridmetrics_table)
print(dt_gridmetrics_table[ , .(Col, Row, elev_mean, CanopyCover, stratum_02_mean, stratum_02_count)])

#4. build a quick preview plot of the canopy height model
#  - FUSION2's GeoTIFFs carry the tile's coordinate system and extent, so
#    x and y here are the tile's true projected coordinates (EPSG:2927, ft)
r_chm = terra::rast(path_chm)
df_chm = as.data.frame(r_chm, xy = TRUE)
names(df_chm) = c("x", "y", "canopy_height_ft")

p_chm = ggplot2::ggplot(df_chm, ggplot2::aes(x = x, y = y, fill = canopy_height_ft)) +
  ggplot2::geom_raster() +
  ggplot2::scale_fill_viridis_c(name = "Canopy height (ft)") +
  ggplot2::coord_equal() +
  ggplot2::labs(
      title = "Example canopy height model"
    , subtitle = "Lidar tile TC_1372 -- 200 x 200 ft closed-canopy forest clip"
    , x = "Easting (ft, EPSG:2927)"
    , y = "Northing (ft, EPSG:2927)"
    ) +
  ggplot2::theme_minimal(base_size = 12)

#5. save the preview figure through the standard versioned-figure path and
#   display it immediately
path_chm_preview = RForInvt::file_version(
    file.path(dir_derived, "chm_preview.tif")
  , increment = TRUE
  , note = "quick-look preview plot of the example canopy height model"
  )
ggplot2::ggsave(path_chm_preview, p_chm, device = "tiff", width = 6, height = 5)
shell.exec(normalizePath(path_chm_preview))
```

# Step 6: Compute per-plot summary and height-strata metrics with `cloudmetrics`

`cloudmetrics` computes statistical summaries (elevation percentiles, canopy cover, canopy relief ratio, and standard deviations) for a point cloud file or directory of files, with optional height normalization against a ground DTM raster or directory of DTM tiles.

Like `gridmetrics` in Step 3, `/nodata:` and `/noheight:` are named
explicitly below at their defaults: `/nodata:NA` would mark the whole
output row `NA` if the clip had zero points at all, and `/noheight:0` marks
the height-dependent columns (the `elev_*`/`int_*` statistic bundle, and
each stratum bucket's mean/stddev/min/max) `0` if points exist but none
clear `/minht:2.0` (or don't land in that bucket) -- neither condition
applies to this particular clip (it has plenty of canopy returns), but the
columns are always present and worth naming for a reader comparing this
output against a clip that does hit one of those cases. `/strata` below
uses gridmetrics' Step 3 thresholds, bucketing this clip's points into the
same eight height strata and reporting `cloudmetrics`' own per-bucket
count/proportion/mean/stddev/min/max summary as CSV columns -- there is no
raster to append bands to here, since `cloudmetrics` produces one summary
row (or one row per `/shape` polygon feature), not a grid.

```{r}
#| label: step-6-cloudmetrics

#1. run cloudmetrics on the example LAS file with ground DTM height
#   normalization and the same height strata thresholds gridmetrics used
#   in Step 3
path_cloudmetrics_out = RForInvt::file_version(
    file.path(dir_tables, "cloudmetrics_normalized.csv")
  , increment = TRUE
  , note = "cloudmetrics summary metrics with ground DTM normalization and height strata"
  )

fn_run_fusion_tool(
    path_cloudmetrics
  , args = c(
        path_las_example
      , paste0("/ground:", path_ground_dem)
      , paste0("/output:", path_cloudmetrics_out)
      , "/minht:2.0"
      , "/nodata:NA"
      , "/noheight:0"
      , "/strata:0.5,2.0,5.0,10.0,20.0,50.0,100.0"
      )
  , tool_label = "cloudmetrics"
  )

#2. read back and display the computed metrics -- cloudmetrics now writes
#   the same full elevation/intensity statistic bundle gridmetrics does
#   (AAD, MAD-median/mode, L-moments, quadratic/cubic mean, percentiles,
#   ...), not just the handful of columns earlier releases wrote
dt_cloudmetrics = data.table::fread(path_cloudmetrics_out)
print(dt_cloudmetrics)

#3. spot-check the height-strata columns specifically -- same bucket
#   layout and naming convention as gridmetrics' /strata columns (Step 3),
#   just one summary row instead of one row per grid cell
nms_strata_cols = grep("^stratum_", names(dt_cloudmetrics), value = TRUE)
print(dt_cloudmetrics[ , ..nms_strata_cols])
```

# Session Info

```{r}
#| label: session-info

sessionInfo()
```
