---
title: "Example FUSION2 Lidar Pipeline: Ground Filter, Canopy Model, Grid Metrics"
author: "Jacob Strunk"
date: today
format: html
execute:
  echo: true
  warning: false
  message: false
---

<!--
Pipeline script -- runs an ordered sequence of FUSION2 command-line tools
against a small example lidar point cloud tile. Not a scientific finding, so
this follows the pipeline template (execution-ordered "# Step N" headings),
not the manuscript-style analysis template. See
r_coding_standards.md
Section 19.
-->

# Setup

## Reset working directory to the project root

<!--
This script lives in R/analysis/, but Quarto executes chunks with the
working directory set to the .qmd's own folder by default -- so
project-relative paths like "data/raw" would otherwise resolve against
R/analysis/, not the project root. root.dir only takes effect starting
with the chunk after it is set.
-->

```{r}
#| label: root-dir

knitr::opts_knit$set(root.dir = "../..")
```

## Load libraries and set default options

```{r}
#| label: setup

# data.table for the small metrics summary table; terra for reading back
# and previewing the GeoTIFF rasters the FUSION2 tools write; RForInvt for
# output versioning; ggplot2 for the one preview figure.
library(data.table)
library(terra)
library(RForInvt)
library(ggplot2)

knitr::opts_chunk$set(
    echo      = TRUE
  , warning   = FALSE
  , message   = FALSE
  )

options(scipen = 10E6)

# Directories used throughout this pipeline.
dir_lidar_raw = file.path("data", "raw", "lidar")
dir_derived   = file.path("data", "derived", "lidar_example")
dir_tables    = file.path("output", "tables")
dir.create(dir_derived, recursive = TRUE, showWarnings = FALSE)
dir.create(dir_tables, recursive = TRUE, showWarnings = FALSE)

# The compiled FUSION2 CLI tools live in the sibling FUSION2 repo's bin/
# folder -- this example repo calls the already-built executables, it does
# not build them.
dir_fusion_bin = "<FUSION2 bin directory>"
path_groundfilter = file.path(dir_fusion_bin, "groundfilter.exe")
path_canopymodel  = file.path(dir_fusion_bin, "canopymodel.exe")
path_gridmetrics  = file.path(dir_fusion_bin, "gridmetrics.exe")
path_cloudmetrics = file.path(dir_fusion_bin, "cloudmetrics.exe")

# The small example point cloud tile this pipeline runs against. See the
# "Pipeline overview" section below for how it was sourced and reduced.
path_las_example = file.path(
    dir_lidar_raw
  , "USGS_LPC_WA_6County_A24_w2049750n409500_clip200ft_las12.las"
  )
if(!file.exists(path_las_example)) stop("Example lidar tile not found at ", path_las_example)
```

## Define helper functions used throughout this pipeline

```{r}
#| label: helper-functions

#1. define a helper that runs one FUSION2 CLI tool and reports what happened
#  - wraps system2() in try() per r_coding_standards.md Section 16 (no
#    tryCatch())
#  - returns the captured console output invisibly, and stops with a clear
#    message if the tool exits with a non-zero status
fn_run_fusion_tool = function(exe_path, args, tool_label){
  if(!file.exists(exe_path)) stop("FUSION2 tool not found: ", exe_path)
  result = try(
      system2(exe_path, args = args, stdout = TRUE, stderr = TRUE)
    , silent = TRUE
    )
  exit_status = attr(result, "status")
  has_error   = inherits(result, "try-error") || (!is.null(exit_status) && exit_status != 0)
  cat(paste(result, collapse = "\n"), "\n")
  if(has_error) stop(tool_label, " exited with an error -- see console output above.")
  invisible(result)
  }

#2. define a helper that summarizes one GeoTIFF raster's cell values into a
#   single-row data.table -- reused across the three rasters this pipeline
#   produces
fn_summarize_raster = function(path_raster, layer_name){
  r = terra::rast(path_raster)
  v = terra::values(r[[layer_name]])
  data.table::data.table(
      layer   = layer_name
    , n_cells = length(v)
    , n_na    = sum(is.na(v))
    , min_val = min(v, na.rm = TRUE)
    , mean_val = mean(v, na.rm = TRUE)
    , max_val = max(v, na.rm = TRUE)
    )
  }
```

# Pipeline overview

This pipeline demonstrates a standard ground-classification-to-metrics
lidar workflow using three FUSION2 command-line tools -- `groundfilter`,
`canopymodel`, and `gridmetrics` -- against one small, real example point
cloud tile.

**Example tile source.** The input tile is a small clip of a public USGS
3DEP lidar point cloud, project `WA_6County_A24`, tile
`w2049750n409500`, published 2026-07-02 at
`https://rockyweb.usgs.gov/vdelivery/Datasets/Staged/Elevation/LPC/Projects/WA_6County_A24/WA_6County_1_A24/LAZ/USGS_LPC_WA_6County_A24_w2049750n409500.laz`.
USGS 3DEP lidar point clouds are a U.S. government work product and are in
the public domain. The full tile (2250 x 2250 ft, roughly 6.1 million
points, 23.8 MB compressed) was clipped down to a 200 x 200 ft area near
the tile center (51,737 points) and converted from its native LAS 1.4
point format 6 to LAS 1.2 point format 0, then committed to
`data/raw/lidar/` at roughly 1 MB. The projected coordinate system is
NAD83(HARN) / Washington South (ftUS), EPSG:2927, vertical datum NAVD88
(ftUS, Geoid18).

**Point format note.** The current FUSION2 build's point-cloud readers
return empty results (zero points, all-NA output rasters) on this tile's
native LAZ-compressed, LAS 1.4 point format 6 encoding, but read the same
points correctly once decompressed and re-encoded as LAS 1.2 point format
0 -- confirmed against this same build's own `test_output/` artifacts,
which show the identical all-NA pattern on a point format 6 input. The
committed example tile therefore ships in LAS 1.2 point format 0 so this
pipeline runs cleanly against the current build; no FUSION2 source code was
changed to arrive at this workaround.

Steps:

1. Classify ground points and build a ground surface DEM (`groundfilter`).
2. Build a canopy height model from the point cloud and ground DEM
   (`canopymodel`).
3. Compute a grid of lidar elevation and canopy metrics (`gridmetrics`).
4. Summarize the three derived rasters into one small metrics table and
   preview the canopy height model.

# Step 1: Classify ground points and build a ground surface DEM

`groundfilter` separates ground returns from the rest of the point cloud
and interpolates them into a ground surface raster, which the next two
steps use to normalize point and canopy heights above ground.

```{r}
#| label: step-1-groundfilter

#1. mint a versioned output path for the ground DEM
path_ground_dem = RForInvt::file_version(
    file.path(dir_derived, "ground_dem.tif")
  , increment = TRUE
  , note = "ground DEM from example 200x200 ft lidar clip"
  )

#2. run groundfilter against the example tile
#  - the CLI-harmonization pass renamed this flag from /output-dem: to
#    /output-raster:; the old name is still silently accepted (exit 0)
#    but produces no output file at all, so this was previously a silent
#    no-op -- confirmed by direct testing against the current build
fn_run_fusion_tool(
    path_groundfilter
  , args = c(
        path_las_example
      , paste0("/output-raster:", path_ground_dem)
      , "/cellsize:5"
      )
  , tool_label = "groundfilter"
  )
```

# Step 2: Build a canopy height model

`canopymodel` builds a top-of-canopy surface from the point cloud and
normalizes it to height above ground using the DEM from Step 1.

```{r}
#| label: step-2-canopymodel

#1. mint a versioned output path for the canopy height model
path_chm = RForInvt::file_version(
    file.path(dir_derived, "chm.tif")
  , increment = TRUE
  , note = "canopy height model from example 200x200 ft lidar clip"
  )

#2. run canopymodel against the example tile, using Step 1's ground DEM
fn_run_fusion_tool(
    path_canopymodel
  , args = c(
        path_las_example
      , paste0("/output:", path_chm)
      , paste0("/ground:", path_ground_dem)
      , "/cellsize:5"
      )
  , tool_label = "canopymodel"
  )
```

# Step 3: Compute a grid of lidar elevation and canopy metrics

`gridmetrics` computes a multi-band raster of standard lidar metrics
(mean and standard deviation of elevation, the 95th height percentile,
canopy cover, and point density) over the example tile, again normalized
using Step 1's ground DEM.

`/nodata:` and `/noheight:` are named explicitly below even though both are
left at their defaults, so a reader sees them: `/nodata:NA` marks a grid
cell with zero lidar returns at all (a genuine data gap -- nothing landed
there), while `/noheight:0` marks a cell that has returns but none clear
the `/minht` cutoff (a real "no canopy here" answer, not missing data).
`terra::rast()` already treats a GeoTIFF's registered NoData value
(including `NaN`, which `/nodata:NA` resolves to) as `NA` on read, and
`data.table::fread()`'s default `na.strings` already includes `"NA"` for
the CSV summary table -- both example-script readers below pick up the
convention automatically, no reader-side code change needed.

```{r}
#| label: step-3-gridmetrics

#1. gridmetrics writes its own output filename into outdir, based on the
#   input file's name -- point it at a dedicated subfolder so the result
#   is easy to find and version
dir_gridmetrics_out = file.path(dir_derived, "gridmetrics")
dir.create(dir_gridmetrics_out, recursive = TRUE, showWarnings = FALSE)

#2. run gridmetrics against the example tile
fn_run_fusion_tool(
    path_gridmetrics
  , args = c(
        path_las_example
      , paste0("/ground:", path_ground_dem)
      , "/cellsize:10"
      , paste0("/outdir:", dir_gridmetrics_out)
      , "/nodata:NA"
      , "/noheight:0"
      )
  , tool_label = "gridmetrics"
  )

#3. mint a versioned copy of the resulting raster at a stable name
path_gridmetrics_raw = list.files(
    dir_gridmetrics_out
  , pattern = "_gridmetrics\\.tif$"
  , full.names = TRUE
  )[1]
path_gridmetrics = RForInvt::file_version(
    file.path(dir_derived, "gridmetrics.tif")
  , increment = TRUE
  , note = "gridmetrics raster from example 200x200 ft lidar clip"
  )
file.copy(path_gridmetrics_raw, path_gridmetrics, overwrite = TRUE)
```

# Step 4: Summarize the derived rasters and preview the canopy height model

This step confirms the three rasters carry real, non-empty values (not
just placeholder output), then saves a small summary table and a quick
preview figure of the canopy height model.

```{r}
#| label: step-4-summarize

#1. summarize each raster's cell values into one small comparison table
dt_summary = data.table::rbindlist(list(
    fn_summarize_raster(path_ground_dem, "ground_elevation")
  , fn_summarize_raster(path_chm, "canopy_height")
  , fn_summarize_raster(path_gridmetrics, "elev_mean")
  , fn_summarize_raster(path_gridmetrics, "canopy_cover")
  , fn_summarize_raster(path_gridmetrics, "point_density")
  # elev_aad (average absolute deviation) and r1count (first-return count)
  # are part of gridmetrics' full elevation-statistic bundle and
  # per-return-number counts -- summarized here to confirm those bands
  # exist and carry real values, not just the handful of bands already
  # checked above
  , fn_summarize_raster(path_gridmetrics, "elev_aad")
  , fn_summarize_raster(path_gridmetrics, "r1count")
  ))
print(dt_summary)

#2. save the summary table as a small, tracked output
path_summary_table = RForInvt::file_version(
    file.path(dir_tables, "lidar_example_raster_summary.csv")
  , increment = TRUE
  , note = "raster summary stats for the example lidar pipeline"
  )
data.table::fwrite(dt_summary, path_summary_table)

#3. build a quick preview plot of the canopy height model
#  - the current FUSION2 build does not attach a coordinate reference or
#    real-world extent to its GeoTIFF output (cell indices only), so this
#    preview uses relative grid coordinates rather than the tile's true
#    projected coordinates
r_chm = terra::rast(path_chm)
df_chm = as.data.frame(r_chm, xy = TRUE)
names(df_chm) = c("x", "y", "canopy_height_ft")

p_chm = ggplot2::ggplot(df_chm, ggplot2::aes(x = x, y = y, fill = canopy_height_ft)) +
  ggplot2::geom_raster() +
  ggplot2::scale_fill_viridis_c(name = "Canopy height (ft)") +
  ggplot2::coord_equal() +
  ggplot2::labs(
      title = "Example canopy height model"
    , subtitle = "USGS 3DEP tile WA_6County_A24, w2049750n409500 -- 200 x 200 ft clip"
    , x = "Grid column (relative)"
    , y = "Grid row (relative)"
    ) +
  ggplot2::theme_minimal(base_size = 12)

#4. save the preview figure through the standard versioned-figure path and
#   display it immediately
path_chm_preview = RForInvt::file_version(
    file.path(dir_derived, "chm_preview.tif")
  , increment = TRUE
  , note = "quick-look preview plot of the example canopy height model"
  )
ggplot2::ggsave(path_chm_preview, p_chm, device = "tiff", width = 6, height = 5)
shell.exec(normalizePath(path_chm_preview))
```

# Step 5: Compute summary plot and clip metrics with `cloudmetrics`

`cloudmetrics` computes statistical summaries (elevation percentiles, canopy cover, canopy relief ratio, and standard deviations) for a point cloud file or directory of files, with optional height normalization against a ground DTM raster or directory of DTM tiles.

Like `gridmetrics` in Step 3, `/nodata:` and `/noheight:` are named
explicitly below at their defaults: `/nodata:NA` would mark the whole
output row `NA` if the clip had zero points at all, and `/noheight:0` marks
the height-dependent columns (the `elev_*`/`int_*` statistic bundle) `0`
if points exist but none clear `/minht:2.0` -- neither condition applies to
this particular clip (it has plenty of canopy returns), but the columns
are always present and worth naming for a reader comparing this output
against a clip that does hit one of those cases.

```{r}
#| label: step-5-cloudmetrics

#1. run cloudmetrics on the example LAS file with ground DTM height normalization
path_cloudmetrics_out = RForInvt::file_version(
    file.path(dir_tables, "cloudmetrics_normalized.csv")
  , increment = TRUE
  , note = "cloudmetrics summary metrics with ground DTM normalization"
  )

fn_run_fusion_tool(
    path_cloudmetrics
  , args = c(
        path_las_example
      , paste0("/ground:", path_ground_dem)
      , paste0("/output:", path_cloudmetrics_out)
      , "/minht:2.0"
      , "/nodata:NA"
      , "/noheight:0"
      )
  , tool_label = "cloudmetrics"
  )

#2. read back and display the computed metrics -- cloudmetrics now writes
#   the same full elevation/intensity statistic bundle gridmetrics does
#   (AAD, MAD-median/mode, L-moments, quadratic/cubic mean, percentiles,
#   ...), not just the handful of columns earlier releases wrote
dt_cloudmetrics = data.table::fread(path_cloudmetrics_out)
print(dt_cloudmetrics)
```

# Session Info

```{r}
#| label: session-info

sessionInfo()
```
