---
title: "Example WA DNR RSFRIS / SFIS Plot-Tree Join"
author: "Jacob Strunk"
date: today
format: html
execute:
  echo: true
  warning: false
  message: false
---

<!--
Pipeline script -- builds a small, non-confidential example subset of two
Washington DNR forest inventory programs (RSFRIS and SFIS) and demonstrates
a plot-tree join and summary. Not a scientific finding, so this follows the
pipeline template (execution-ordered "# Step N" headings), not the
manuscript-style analysis template. See
r_coding_standards.md
Section 19.

This is independent of 001_example_gridmetrics_lidar_pipeline.qmd -- the WA
DNR plots used here are not known to overlap the example lidar tile, so
these are two separate, independently-runnable examples rather than one
combined lidar-plus-plots workflow.
-->

# Setup

## Reset working directory to the project root

```{r}
#| label: root-dir

knitr::opts_knit$set(root.dir = "../..")
```

## Load libraries and set default options

```{r}
#| label: setup

library(data.table)
library(RForInvt)

knitr::opts_chunk$set(
    echo      = TRUE
  , warning   = FALSE
  , message   = FALSE
  )

options(scipen = 10E6)

# The true, full-size WA DNR source tables live outside this repo -- this
# example only ever writes a small subset of them into data/raw/dnr/, which
# is what steps 3+ actually read. Resolve dynamically via Box/sync or environment variables.
dir_box_sync = if (nzchar(Sys.getenv("BOX_PATH")) && dir.exists(Sys.getenv("BOX_PATH"))) {
  normalizePath(Sys.getenv("BOX_PATH"), winslash = "/")
} else if (nzchar(Sys.getenv("USERPROFILE")) && dir.exists(file.path(Sys.getenv("USERPROFILE"), "Box", "sync"))) {
  normalizePath(file.path(Sys.getenv("USERPROFILE"), "Box", "sync"), winslash = "/")
} else {
  normalizePath(file.path("~", "Box", "sync"), winslash = "/", mustWork = FALSE)
}
dir_dnr_source = file.path(dir_box_sync, "data", "WA", "DNR")
dir_dnr_subset = file.path("data", "raw", "dnr")
dir_tables     = file.path("output", "tables")
dir.create(dir_dnr_subset, recursive = TRUE, showWarnings = FALSE)
dir.create(dir_tables, recursive = TRUE, showWarnings = FALSE)

n_example_plots = 15
```

## Define helper functions used throughout this pipeline

```{r}
#| label: helper-functions

#1. define a helper that summarizes a tree-level table into one row per
#   plot -- reused for both the RSFRIS and SFIS tree tables below
fn_summarize_trees_by_plot = function(dt_tree, plot_id_col, species_col){
  dt_tree_i = data.table::copy(dt_tree)
  data.table::setnames(dt_tree_i, c(plot_id_col, species_col), c("plot_id", "species"))
  dt_summary = dt_tree_i[
      , .(
            n_trees = .N
          , n_species = data.table::uniqueN(species)
          , top_species = names(sort(table(species), decreasing = TRUE))[1]
          )
      , by = plot_id
      ]
  dt_summary
  }
```

# Pipeline overview

This pipeline builds a small example subset of two Washington State
Department of Natural Resources (WA DNR) forest inventory programs --
RSFRIS (Remote Sensing Forest Resource Inventory System) and SFIS (Stand-
level Forest Inventory System) -- then demonstrates a plot-level to
tree-level join within each program.

Steps:

1. Screen the source tables for coordinate fields and record which tables
   are safe to use in this public example repo.
2. Build a small (`r n_example_plots`-plot) example subset of the
   non-coordinate RSFRIS and SFIS tables and commit it to
   `data/raw/dnr/`.
3. Load the committed subset and join each program's tree table to its
   plot-level metrics table, summarizing tree counts and species by plot.
4. Combine both programs' plot-level summaries into one comparison table.

# Step 1: Screen source tables for coordinate fields

WA DNR plot identifiers in both programs are UUIDs with no public lookup
back to a real-world location, which is different from the FIA true-
coordinate case this project's standing guidance otherwise treats as a
hard never-commit rule (see `git_practices.md` Section 8). That said, two
of the source tables carry exact projected and/or geographic coordinates
tied to an identifiable plot or sample point, and are excluded from this
repo on that basis alone, independent of the UUID/no-lookup point above:

- `RSFRIS_20260819/rsfris_plot_metadata.rds` carries `X`/`Y` projected
  coordinates keyed by `PLOT_ID` and a human-readable `PLOT_NAME`.
- `sfis_20250218/shp/SL_SFIS_SAMPLE_POINT.shp` carries `X`/`Y` (state
  plane feet) and `LON`/`LAT` (decimal degrees, roughly 5 decimal places)
  keyed by `POINT_ID` and `POINT_NM`.

Neither file is read anywhere in this pipeline, and neither is committed
to this repo. Everything used below comes from the three RSFRIS/SFIS
attribute tables confirmed to carry no coordinate columns:
`rsfris_plot_forest_metrics.rds`, `rsfris_tree.rds`,
`sfis_plot_data_20250218.csv`, and `sfis_tree_data_20250218.csv`.

```{r}
#| label: step-1-screen-coordinates

#1. list the non-coordinate source tables this pipeline is allowed to read
path_rsfris_metrics = file.path(dir_dnr_source, "RSFRIS_20260819", "rsfris_plot_forest_metrics.rds")
path_rsfris_tree     = file.path(dir_dnr_source, "RSFRIS_20260819", "rsfris_tree.rds")
path_sfis_plot       = file.path(dir_dnr_source, "sfis_20250218", "sfis_plot_data_20250218.csv")
path_sfis_tree       = file.path(dir_dnr_source, "sfis_20250218", "sfis_tree_data_20250218.csv")

#2. guard against ever reading the excluded, coordinate-bearing files by
#   name -- this is a documentation safeguard, not a substitute for the
#   screening judgment above
nms_excluded = c("rsfris_plot_metadata.rds", "SL_SFIS_SAMPLE_POINT.shp")
cat("Excluded (carry exact coordinates):", paste(nms_excluded, collapse = ", "), "\n")
```

# Step 2: Build a small example subset of the RSFRIS and SFIS tables

This step samples `r n_example_plots` plots from each program (present in
both that program's plot-level and tree-level table) and writes the
subset to `data/raw/dnr/`. It only runs if the subset files are not
already present -- this is a fixed demo dataset meant to stay stable
across runs, not a rebuild-on-every-render cache in the sense of
`r_coding_standards.md` Section 13, since the source tables it reads live
outside this repo and are not expected to change.

```{r}
#| label: step-2-build-subset

path_rsfris_metrics_sub = file.path(dir_dnr_subset, "rsfris_plot_forest_metrics_subset.csv")
path_rsfris_tree_sub    = file.path(dir_dnr_subset, "rsfris_tree_subset.csv")
path_sfis_plot_sub      = file.path(dir_dnr_subset, "sfis_plot_data_subset.csv")
path_sfis_tree_sub      = file.path(dir_dnr_subset, "sfis_tree_data_subset.csv")

nms_subset_paths = c(path_rsfris_metrics_sub, path_rsfris_tree_sub, path_sfis_plot_sub, path_sfis_tree_sub)

if(!all(file.exists(nms_subset_paths))){

  set.seed(42)

  #1. load the full RSFRIS non-coordinate tables and sample common plots
  dt_rsfris_metrics_all = data.table::setDT(readRDS(path_rsfris_metrics))
  dt_rsfris_tree_all    = data.table::setDT(readRDS(path_rsfris_tree))
  ids_rsfris_common = intersect(dt_rsfris_metrics_all$PLOT_ID, dt_rsfris_tree_all$PLOT_ID)
  ids_rsfris_sample = sample(ids_rsfris_common, size = n_example_plots)

  #2. subset and write the RSFRIS example tables
  dt_rsfris_metrics_sub = dt_rsfris_metrics_all[PLOT_ID %in% ids_rsfris_sample]
  dt_rsfris_tree_sub    = dt_rsfris_tree_all[PLOT_ID %in% ids_rsfris_sample]
  data.table::fwrite(dt_rsfris_metrics_sub, path_rsfris_metrics_sub)
  data.table::fwrite(dt_rsfris_tree_sub, path_rsfris_tree_sub)

  #3. load the full SFIS non-coordinate tables and sample common plots
  dt_sfis_plot_all = data.table::fread(path_sfis_plot)
  dt_sfis_tree_all  = data.table::fread(path_sfis_tree)
  ids_sfis_common = intersect(dt_sfis_plot_all$PLOT_ID, dt_sfis_tree_all$PLOT_ID)
  ids_sfis_sample = sample(ids_sfis_common, size = n_example_plots)

  #4. subset and write the SFIS example tables
  dt_sfis_plot_sub = dt_sfis_plot_all[PLOT_ID %in% ids_sfis_sample]
  dt_sfis_tree_sub  = dt_sfis_tree_all[PLOT_ID %in% ids_sfis_sample]
  data.table::fwrite(dt_sfis_plot_sub, path_sfis_plot_sub)
  data.table::fwrite(dt_sfis_tree_sub, path_sfis_tree_sub)

  } else {

  cat("Example subset already present under", dir_dnr_subset, "-- skipping rebuild.\n")

  }
```

# Step 3: Join each program's tree table to its plot-level metrics

Each program's tree table is summarized to one row per plot (tree count,
species richness, most common species) using the shared
`fn_summarize_trees_by_plot()` helper, then left-joined onto that
program's plot-level metrics table by `plot_id`.

```{r}
#| label: step-3-join

#1. load the committed example subset
dt_rsfris_metrics = data.table::fread(path_rsfris_metrics_sub)
dt_rsfris_tree    = data.table::fread(path_rsfris_tree_sub)
dt_sfis_plot      = data.table::fread(path_sfis_plot_sub)
dt_sfis_tree      = data.table::fread(path_sfis_tree_sub)

#2. summarize each program's tree table by plot
dt_rsfris_tree_summary = fn_summarize_trees_by_plot(dt_rsfris_tree, "PLOT_ID", "SPECIES")
dt_sfis_tree_summary   = fn_summarize_trees_by_plot(dt_sfis_tree, "PLOT_ID", "SPECIES")

#3. join the tree summary onto each program's plot-level metrics table
dt_rsfris_joined = merge(
    dt_rsfris_metrics
  , dt_rsfris_tree_summary
  , by.x = "PLOT_ID"
  , by.y = "plot_id"
  , all.x = TRUE
  )
dt_sfis_joined = merge(
    dt_sfis_plot
  , dt_sfis_tree_summary
  , by.x = "PLOT_ID"
  , by.y = "plot_id"
  , all.x = TRUE
  )

cat("RSFRIS joined plots:", nrow(dt_rsfris_joined), "\n")
cat("SFIS joined plots:", nrow(dt_sfis_joined), "\n")
```

# Step 4: Combine both programs' plot-level summaries

Both programs share the same core stand-metric naming (`BA`, `HTMAX`,
`HT_T100`, `QMD`), so this step selects those columns plus the tree-join
results from each program's joined table and stacks them into one small
comparison table.

```{r}
#| label: step-4-combine

#1. select the common comparison columns from each program's joined table
nms_common = c("PLOT_ID", "n_trees", "n_species", "top_species", "BA", "HTMAX", "HT_T100", "QMD")
dt_rsfris_common = dt_rsfris_joined[ , ..nms_common]
dt_rsfris_common[ , source := "RSFRIS"]
dt_sfis_common = dt_sfis_joined[ , ..nms_common]
dt_sfis_common[ , source := "SFIS"]

#2. combine into one comparison table
dt_combined = data.table::rbindlist(list(dt_rsfris_common, dt_sfis_common))
data.table::setcolorder(dt_combined, c("source", "PLOT_ID", "n_trees", "n_species", "top_species", "BA", "HTMAX", "HT_T100", "QMD"))
print(dt_combined)

#3. save the combined comparison table as a small, tracked output
path_combined_table = RForInvt::file_version(
    file.path(dir_tables, "dnr_example_plot_summary.csv")
  , increment = TRUE
  , note = "example RSFRIS/SFIS plot-tree join and summary"
  )
data.table::fwrite(dt_combined, path_combined_table)
```

# Session Info

```{r}
#| label: session-info

sessionInfo()
```
