---
title: "Advanced FUSION2 Pipeline Orchestration: Resuming, Retrying, and Rebuilding"
author: "Jacob Strunk"
date: today
format: html
execute:
  echo: true
  warning: false
  message: false
---

<!--
Pipeline script -- continues from 003_example_pipeline_orchestrator.qmd's
Step 5 (/tool:groundfilter) output directory to demonstrate pipeline.exe's
resume/idempotency behavior, /retryfailed, /rebuild, and its current
inability to start a chain from a pre-built raster instead of a point
cloud. Follows the pipeline template (execution-ordered "# Step N"
headings), not the manuscript-style analysis template.

Run 003_example_pipeline_orchestrator.qmd first -- Step 3 below reruns its
Step 5 command verbatim against the same output directory, so this script
is picking up real, already-recorded state rather than starting cold.
-->

# Setup

## Reset working directory to the project root

```{r}
#| label: root-dir

knitr::opts_knit$set(root.dir = "../..")
```

## Load libraries and set default options

```{r}
#| label: setup

# data.table for state-manifest parsing and summary tables; RForInvt for
# output versioning. No raster-reading package is needed here -- every
# step in this script inspects the state manifest and console output
# directly rather than the rasters themselves (003 already covers that).
library(data.table)
library(RForInvt)

knitr::opts_chunk$set(
    echo      = TRUE
  , warning   = FALSE
  , message   = FALSE
  )

options(scipen = 10E6)

# Same derived-data folder 003 used -- this script's resume/retry/rebuild
# steps depend on genuinely picking up the state 003's Step 5 left behind.
dir_lidar_raw   = file.path("data", "raw", "lidar")
dir_dtm_raw     = file.path("data", "raw", "lidar_dtm_examples")
dir_derived     = file.path("data", "derived", "pipeline_example")
dir_tables      = file.path("output", "tables")
dir_single_out  = file.path(dir_derived, "single_tool_out")
dir.create(dir_derived, recursive = TRUE, showWarnings = FALSE)
dir.create(dir_tables, recursive = TRUE, showWarnings = FALSE)

# The compiled FUSION2 CLI tools are not bundled with this repo -- build
# FUSION2 from source (https://github.com/jstrunk001/FUSION2) and either put
# its bin/ folder on PATH or point the FUSION2_BIN_DIR environment variable
# at it.
fn_find_fusion_tool = function(tool_name) {
  path_on_path = Sys.which(tool_name)
  if (nzchar(path_on_path)) return(normalizePath(path_on_path, winslash = "/"))
  dir_env = Sys.getenv("FUSION2_BIN_DIR")
  if (nzchar(dir_env)) {
    path_in_env_dir = file.path(dir_env, tool_name)
    if (file.exists(path_in_env_dir)) return(normalizePath(path_in_env_dir, winslash = "/"))
  }
  stop(
      "Could not find ", tool_name, ". Build FUSION2 from source "
    , "(https://github.com/jstrunk001/FUSION2) and either add its bin/ "
    , "folder to PATH or set the FUSION2_BIN_DIR environment variable to it."
    )
}
path_pipeline  = fn_find_fusion_tool("pipeline.exe")
path_catalog   = fn_find_fusion_tool("catalog.exe")

# The same small example point cloud tile used throughout this repo.
path_las_example = file.path(
    dir_lidar_raw
  , "TC_1372_forest200ft.laz"
  )
if(!file.exists(path_las_example)) stop("Example lidar tile not found at ", path_las_example)
```

## Define helper functions used throughout this pipeline

```{r}
#| label: helper-functions

#1. define a helper that runs one already-built, literal command-line
#   string and reports what happened, without stopping the script on
#   failure -- Step 7's external-DTM rejection is itself the finding, not
#   something to hide, so this helper returns a result list for the
#   caller to inspect instead of calling stop()
#  - the command is printed exactly as a person would type it at a real
#    terminal; internally it is run with "2>&1" appended (so a rejection
#    message on stderr, like Step 7 below, is actually captured) and
#    wrapped in one extra pair of quotes before being handed to shell(),
#    which is a documented cmd.exe quirk (cmd /c strips a single
#    leading/trailing quote pair from a command line that contains more
#    than one quoted segment) -- both are R/Windows plumbing details, not
#    part of pipeline.exe's own syntax, so they are kept out of the
#    printed command text entirely
fn_run_cmd = function(cmd, run_label){
  cat(run_label, "-- command:\n  ", cmd, "\n\n", sep = "")
  result = try(shell(paste0('"', cmd, ' 2>&1', '"'), intern = TRUE), silent = TRUE)
  exit_status = attr(result, "status")
  exit_status = if(is.null(exit_status)) 0L else exit_status
  console_text = if(inherits(result, "try-error")) as.character(result) else result
  cat(run_label, "-- console output:\n")
  cat(paste(console_text, collapse = "\n"), "\n\n")
  list(
      label     = run_label
    , exit_code = exit_status
    , succeeded = identical(exit_status, 0L)
    , console   = console_text
    )
  }

#2. define a helper that reads a pipeline.exe run's state manifest
#   (_processing/pipeline_state.csv) into a data.table, returning an empty
#   0-row table with the expected columns if the run never got far enough
#   to write one
fn_read_pipeline_state = function(dir_output){
  path_state = file.path(dir_output, "_processing", "pipeline_state.csv")
  state_cols = c("tile", "stage", "status", "output_path", "exit_code", "started_at", "finished_at")
  if(!file.exists(path_state)) return(data.table::data.table(matrix(character(0), ncol = length(state_cols), dimnames = list(NULL, state_cols))))
  data.table::fread(path_state, colClasses = "character")
  }
```

# Pipeline overview

`003_example_pipeline_orchestrator.qmd` covers the shape of every valid
`pipeline.exe` invocation. This script covers what happens when the same
command runs more than once against the same output directory -- the
resume/idempotency behavior, `/retryfailed`, and `/rebuild` -- and one
real gap in what `pipeline.exe` can currently start a chain from.

**Example inputs.** The point-cloud input is the same small (`~0.5 MB`,
200 x 200 ft, 70,532-point) window of closed-canopy forest from lidar
tile `TC_1372` used throughout this repo, in its native LAS 1.4 point
format 6 LAZ encoding. The external-DTM input used in Step 7 is a set of
seven small (`~1 MB` each) real USGS 3DEP bare-earth lidar DTMs, one per
Washington forest-type study site (subalpine Cascades, dry east-side,
west-side plantation, old-growth rainforest, and three others), pulled
from the `2026_Compare_DSMs` analysis project's cached, real (not
synthetic) lidar reference rasters -- see `data/raw/lidar_dtm_examples/`
and its per-site subfolders.

# Step 1: Stage the example inputs into their own input directories

Re-stages the point cloud (`file.copy(..., overwrite = TRUE)` is a no-op
if 003 already staged it) and flattens the seven per-site DTM rasters into
one directory with unique filenames for Step 7.

```{r}
#| label: step-1-stage-inputs

#1. stage the point cloud into its own input directory
dir_las_input = file.path(dir_derived, "las_input")
dir.create(dir_las_input, recursive = TRUE, showWarnings = FALSE)
file.copy(path_las_example, dir_las_input, overwrite = TRUE)

#2. flatten the seven per-site DTM rasters into one directory, renamed by
#   site so pipeline.exe sees seven distinct input files rather than seven
#   files that all share the name "usgs_lidar_dtm.tif"
dir_dtm_input = file.path(dir_derived, "dtm_input")
dir.create(dir_dtm_input, recursive = TRUE, showWarnings = FALSE)
dirs_dtm_site = list.dirs(dir_dtm_raw, full.names = TRUE, recursive = FALSE)
for(dir_site_i in dirs_dtm_site){
  site_id_i = basename(dir_site_i)
  file.copy(
      file.path(dir_site_i, "usgs_lidar_dtm.tif")
    , file.path(dir_dtm_input, paste0(site_id_i, ".tif"))
    , overwrite = TRUE
    )
  }
cat("Staged", length(list.files(dir_dtm_input)), "site DTMs into", dir_dtm_input, "\n")
```

# Step 2: Determine the point cloud's real project extent

Same extent-detection step 003 uses, repeated here so this script can run
on its own.

```{r}
#| label: step-2-real-extent

#1. run catalog.exe against the staged input directory
path_catalog_report = file.path(dir_derived, "catalog_report.csv")
cmd_catalog = sprintf(
    '"%s" "%s" /output:"%s"'
  , path_catalog, dir_las_input, path_catalog_report
  )
result_catalog = fn_run_cmd(cmd_catalog, "Step 2: catalog.exe")
dt_catalog = data.table::fread(path_catalog_report)
print(dt_catalog)

#2. pull the real bounding box out of catalog.exe's report
extent_las = c(
    xmin = dt_catalog$MinX[1], ymin = dt_catalog$MinY[1]
  , xmax = dt_catalog$MaxX[1], ymax = dt_catalog$MaxY[1]
  )
cat("Real extent read from catalog.exe:\n")
print(extent_las)

#3. build the /extent: argument string below expects
arg_extent   = paste0(
    "/extent:", extent_las["xmin"], ",", extent_las["ymin"]
  , ",", extent_las["xmax"], ",", extent_las["ymax"]
  )
arg_tilesize = "/tilesize:300,300"
```

# Step 3: Establish the baseline -- 003's single-tool groundfilter run

This is 003's Step 5 command, unchanged, against the same output
directory. If 003 already ran, this rerun is itself a first live example
of resume behavior (see Step 4 for a version that measures it directly);
if not, this establishes the state Steps 4-6 below build on.

```{r}
#| label: step-3-baseline

dir.create(dir_single_out, recursive = TRUE, showWarnings = FALSE)

cmd_groundfilter_single = sprintf(
    '"%s" /tool:groundfilter /input:"%s" /output:"%s" %s %s /buffer:25 /cellsize:5'
  , path_pipeline, dir_las_input, dir_single_out, arg_extent, arg_tilesize
  )
result_baseline = fn_run_cmd(
    cmd_groundfilter_single
  , "Step 3: baseline (/tool:groundfilter)"
  )
print(fn_read_pipeline_state(dir_single_out))
```

# Step 4: Resume behavior -- rerunning the same command

Per `docs/PIPELINE_GUIDE.md` Section 8, re-running the exact same command
should be safe and cheap: any tile/stage already recorded `done` (with its
output file still present) is skipped. This step reruns Step 3's exact
command unchanged and compares the two state manifests.

```{r}
#| label: step-4-resume

#1. capture the state manifest's finished_at timestamps before the rerun
dt_state_before_rerun = fn_read_pipeline_state(dir_single_out)

#2. rerun the identical command from Step 3
result_rerun = fn_run_cmd(
    cmd_groundfilter_single
  , "Step 4: resume -- identical rerun of Step 3's command"
  )

#3. compare state before/after -- a stage whose recorded status was
#   already "done" and whose finished_at timestamp is unchanged was
#   genuinely skipped rather than silently redone
dt_state_after_rerun = fn_read_pipeline_state(dir_single_out)
dt_compare_rerun = merge(
    dt_state_before_rerun[, .(tile, stage, status_before = status, finished_before = finished_at)]
  , dt_state_after_rerun[, .(tile, stage, status_after = status, finished_after = finished_at)]
  , by = c("tile", "stage"), all = TRUE
  )
print(dt_compare_rerun)
```

# Step 5: `/retryfailed` -- retrying only the tiles/stages that failed

`/retryfailed` processes only tiles with a recorded failed stage, leaving
any already-`done` stages untouched.

```{r}
#| label: step-5-retryfailed

cmd_retryfailed = paste(cmd_groundfilter_single, "/retryfailed")
result_retryfailed = fn_run_cmd(cmd_retryfailed, "Step 5: /retryfailed")
print(fn_read_pipeline_state(dir_single_out))
```

# Step 6: `/rebuild` -- forcing every tile/stage to redo regardless of state

`/rebuild` ignores recorded state entirely and redoes every tile/stage,
regardless of whether it was already marked `done`.

```{r}
#| label: step-6-rebuild

dt_state_before_rebuild = fn_read_pipeline_state(dir_single_out)

cmd_rebuild = paste(cmd_groundfilter_single, "/rebuild")
result_rebuild = fn_run_cmd(cmd_rebuild, "Step 6: /rebuild")

dt_state_after_rebuild = fn_read_pipeline_state(dir_single_out)
dt_compare_rebuild = merge(
    dt_state_before_rebuild[, .(tile, stage, finished_before = finished_at)]
  , dt_state_after_rebuild[, .(tile, stage, finished_after = finished_at)]
  , by = c("tile", "stage"), all = TRUE
  )
cat("A stage genuinely rebuilt under /rebuild has a finished_after timestamp\n")
cat("later than finished_before, even though it was already 'done'.\n")
print(dt_compare_rebuild)
```

# Step 7: Feeding `pipeline.exe` an independently-sourced external DTM

Every scenario in this repo so far starts from a raw point cloud, because
that is currently the only input `pipeline.exe` accepts -- by design, the
first stage in any chain must consume a point cloud
(`docs/PIPELINE_GUIDE.md` Section 9, rule 1). This step demonstrates the
workflow this project actually needs next: running a raster-consuming
tool (`topometrics`, computing slope/aspect) directly against a set of
independently-sourced, already-built DTM rasters -- the seven real USGS
3DEP lidar bare-earth DTMs staged in Step 1 -- with no point cloud
involved at all.

**Known limitation (open defect, not yet fixed, confirmed by direct
testing).** `pipeline.exe` currently rejects this outright:

```
Error: The first pipeline stage must consume a point cloud (gridmetrics,
canopymodel, groundfilter, returndensity, filterdata, or thindata) --
"topometrics" expects a raster as input, and no raster exists yet at the
start of a tile.
```

This is a real gap, not a usage mistake: there is currently no way to
start a `pipeline.exe` chain from a pre-built raster instead of a point
cloud, even for a tool like `topometrics` whose only input is a raster.
The command below is written as the fix is expected to make it work --
`/input:` pointing at a directory of GeoTIFF rasters instead of LAS/LAZ
files, with each raster treated as one pre-built tile (skipping the usual
point-cloud buffered-clip step for that tile) -- so once that support
lands, re-running this exact chunk should turn today's rejection into
seven real slope/aspect rasters, one per Washington forest-type site.

```{r}
#| label: step-7-external-dtm

dir_dtm_topo_out = file.path(dir_derived, "external_dtm_topometrics_out")
dir.create(dir_dtm_topo_out, recursive = TRUE, showWarnings = FALSE)

cmd_external_dtm = sprintf(
    '"%s" /tool:topometrics /input:"%s" /output:"%s" /merge'
  , path_pipeline, dir_dtm_input, dir_dtm_topo_out
  )
result_external_dtm = fn_run_cmd(
    cmd_external_dtm
  , "Step 7: /tool:topometrics against 7 real external lidar DTMs (expected to fail today)"
  )
print(fn_read_pipeline_state(dir_dtm_topo_out))
```

# Step 8: Summarize today's real results

```{r}
#| label: step-8-summary

#1. assemble one row per scenario, checked against real evidence rather
#   than just the process exit code -- an exit code of 0 with no output
#   file present is reported as a failure, not a pass
dt_status = data.table::data.table(
    step = c(
        "3: baseline (/tool:groundfilter)"
      , "4: resume (identical rerun)"
      , "5: /retryfailed"
      , "6: /rebuild"
      , "7: external DTM input (topometrics)"
      )
  , exit_code = c(
        result_baseline$exit_code, result_rerun$exit_code
      , result_retryfailed$exit_code, result_rebuild$exit_code
      , result_external_dtm$exit_code
      )
  , expected_output_present = c(
        file.exists(file.path(dir_single_out, "groundfilter.vrt"))
      , file.exists(file.path(dir_single_out, "groundfilter.vrt"))
      , file.exists(file.path(dir_single_out, "groundfilter.vrt"))
      , file.exists(file.path(dir_single_out, "groundfilter.vrt"))
      , file.exists(file.path(dir_dtm_topo_out, "topometrics.vrt"))
      )
  )
print(dt_status)

#2. save this status table as a small, tracked, versioned output
path_status_table = RForInvt::file_version(
    file.path(dir_tables, "pipeline_orchestration_advanced_status.csv")
  , increment = TRUE
  , note = "real pipeline.exe resume/retryfailed/rebuild/external-DTM run status"
  )
data.table::fwrite(dt_status, path_status_table)
```

# Known Limitations

- **`pipeline.exe` cannot start a chain from a pre-built raster.** Every
  chain must begin with a point-cloud-consuming stage; there is currently
  no way to hand `pipeline.exe` an already-built DTM and run a
  raster-only tool like `topometrics` against it -- confirmed by direct
  testing (Step 7).
- **No spatial validation between the point-cloud demo tile and the
  external DTM examples.** The seven DTMs staged in Step 1 are real
  lidar data from `2026_Compare_DSMs`, but from different Washington
  sites than the small point-cloud tile used in Steps 3-6 -- they are
  included to exercise `pipeline.exe`'s external-raster-input handling
  once it exists, not as a co-located accuracy check against this
  script's own `groundfilter` output.

# Session Info

```{r}
#| label: session-info

sessionInfo()
```
