---
title: "Every FUSION2 Tool on One Example Tile, With Output Checks"
author: "Jacob Strunk"
date: today
format: html
execute:
  echo: true
  warning: false
  message: false
---

<!--
Pipeline script -- runs every one of the 15 FUSION2 command-line tools
against the small example lidar tile, then checks each tool's output
against an independent calculation made in R (lidR/terra/sf). Follows the
pipeline template (execution-ordered "# Step N" headings), not the
manuscript-style analysis template.
-->

# Setup

## Reset working directory to the project root

This script lives in `R/analysis/`, but Quarto runs chunks from the
`.qmd`'s own folder, so project-relative paths need the working directory
moved to the project root. The setting takes effect from the next chunk.

```{r}
#| label: root-dir

knitr::opts_knit$set(root.dir = "../..")
```

## Load libraries and set default options

Loads the packages used for independent checks: `lidR` reads the point
clouds, `terra` and `sf` handle rasters and polygons, and `data.table`
holds the check log.

```{r}
#| label: setup

#1. load packages
#  - lidR also defines a crs() function, so every terra call below is
#    written as terra::fn() to avoid picking up the wrong one
library(data.table)
library(terra)
library(sf)
library(lidR)
library(ggplot2)
library(RForInvt)

knitr::opts_chunk$set(
    echo      = TRUE
  , warning   = FALSE
  , message   = FALSE
  , fig.width  = 7
  , fig.height = 5
  )

options(scipen = 10E6)

#2. set project directories
dir_lidar_raw = file.path("data", "raw", "lidar")
dir_dtm_raw   = file.path("data", "raw", "lidar_dtm_examples")
dir_derived   = file.path("data", "derived", "all_tools_example")
dir_tables    = file.path("output", "tables")
dir.create(dir_derived, recursive = TRUE, showWarnings = FALSE)
dir.create(dir_tables, recursive = TRUE, showWarnings = FALSE)

#3. mint one versioned run folder that every tool writes into
#  - file_version() returns a new versioned path each run; it is used
#    here as a folder, so one run's 30-odd tool outputs stay together
dir_run = RForInvt::file_version(
    file.path(dir_derived, "run")
  , increment = TRUE
  , note = "all-tools demonstration run"
  )
dir.create(dir_run, recursive = TRUE, showWarnings = FALSE)
dir_run = normalizePath(dir_run, winslash = "/")
cat("Tool outputs for this run go to:", dir_run, "\n")

#4. locate the FUSION2 tools
#  - the compiled tools are not bundled with this repo -- build FUSION2
#    from source (https://github.com/jstrunk001/FUSION2) and either put
#    its bin/ folder on PATH or point FUSION2_BIN_DIR at it
fn_find_fusion_tool = function(tool_name){
  path_on_path = Sys.which(tool_name)
  if(nzchar(path_on_path)) return(normalizePath(path_on_path, winslash = "/"))
  dir_env = Sys.getenv("FUSION2_BIN_DIR")
  path_in_env_dir = file.path(dir_env, tool_name)
  if(nzchar(dir_env) && file.exists(path_in_env_dir)) return(normalizePath(path_in_env_dir, winslash = "/"))
  stop(
      "Could not find ", tool_name, ". Build FUSION2 from source "
    , "(https://github.com/jstrunk001/FUSION2) and either add its bin/ "
    , "folder to PATH or set the FUSION2_BIN_DIR environment variable to it."
    )
  }

nms_tools = c(
    "catalog", "filterdata", "thindata", "groundfilter", "clipdata"
  , "canopymodel", "returndensity", "densitymetrics", "gridmetrics"
  , "cloudmetrics", "topometrics", "gridsurfacestats", "canopymaxima"
  , "treeseg", "pipeline"
  )
path_tools = sapply(paste0(nms_tools, ".exe"), fn_find_fusion_tool)
names(path_tools) = nms_tools

#5. record which build is being tested, so the check log can be tied to it
dt_build = data.table::data.table(
    tool     = nms_tools
  , path     = path_tools
  , modified = format(file.info(path_tools)$mtime, "%Y-%m-%d %H:%M")
  )
print(dt_build)

#6. point at the example tile and one external DTM
#  - the tile is a 200 x 200 ft window of closed-canopy forest (92%
#    canopy cover, 95th-percentile height 142 ft) from lidar tile
#    TC_1372 (NV5 Geospatial, 2022), kept in its native LAS 1.4 point
#    format 6 LAZ encoding; see 006 for how the window was chosen.
#    Every clip box, plot, and elevation band below is placed relative to
#    this tile's own extent and elevations, so another tile can be swapped
#    in without editing coordinates
#  - the external DTM is a USGS 3DEP bare-earth raster covering the same
#    ground as the tile (plus a 50 ft pad), with a coordinate system
#    attached; 006 Step 5 fetches it. It is used to check that
#    raster-in/raster-out tools carry the coordinate system through, and
#    its slope and aspect can be compared directly with those from the
#    FUSION2 ground DEM
path_las = normalizePath(file.path(dir_lidar_raw, "TC_1372_forest200ft.laz"), winslash = "/")
path_dtm_external = normalizePath(file.path(dir_dtm_raw, "site_tc_1372_forest", "usgs_3dep_dtm.tif"), winslash = "/")
if(!file.exists(path_las)) stop("Example lidar tile not found at ", path_las)
if(!file.exists(path_dtm_external)) stop("External example DTM not found at ", path_dtm_external)
```

## Define helper functions used throughout this script

Every check in this script goes into one running log instead of stopping
the script. The point of the script is to list every problem in one pass.
Stopping at the first failure would hide the rest. The script stops only
when an output that later steps depend on (the ground DEM or the canopy
height model) is missing entirely.

Each check carries a severity. `hard` means the output is wrong or missing.
`soft` means the output works but something about it deserves a look, such
as a unit label that doesn't match the data or missing coordinate-system
metadata.

```{r}
#| label: helper-functions

#1. create the running check log
#  - an environment is used so helpers can append to it from inside
#    function calls
env_checks = new.env()
env_checks$rows = list()

#2. define a helper that records one check result and prints it
#  - a check whose result can't be computed (NA) counts as a failure,
#    because an output that can't be evaluated isn't a confirmed pass
fn_check = function(step, tool, check, passed, observed = "", expected = "", severity = "hard"){
  passed_in = isTRUE(passed)
  row_i = data.table::data.table(
      step     = step
    , tool     = tool
    , check    = check
    , severity = severity
    , passed   = passed_in
    , observed = paste(format(observed, digits = 8), collapse = ", ")
    , expected = paste(format(expected, digits = 8), collapse = ", ")
    )
  env_checks$rows[[length(env_checks$rows) + 1]] = row_i
  status_label = if(passed_in) "PASS" else if(severity == "hard") "FAIL" else "WARN"
  cat(sprintf("[%s] %s | %s | observed: %s | expected: %s\n", status_label, tool, check, row_i$observed, row_i$expected))
  invisible(passed_in)
  }

#3. define a helper that runs one tool, prints its console output, logs
#   whether it exited cleanly, and records the command it ran
#  - uses try() so a crashing tool is logged, not fatal
#  - the recorded command reads the same on any machine: files this run
#    writes become out/..., and input files become paths relative to the
#    project root (the run folder is replaced first, since it sits inside
#    the project); an argument holding a space is quoted
#  - the FUSION2 user manual shows these recorded commands as its
#    worked examples (see 007_export_fusion2_docs_assets.qmd)
env_checks$commands = list()
dir_project = normalizePath(".", winslash = "/")
fn_portable_args = function(args){
  args_in = gsub(paste0(dir_run, "/"), "out/", args, fixed = TRUE)
  args_in = gsub(dir_run, "out", args_in, fixed = TRUE)
  args_in = gsub(paste0(dir_project, "/"), "", args_in, fixed = TRUE)
  ifelse(grepl(" ", args_in, fixed = TRUE), paste0("\"", args_in, "\""), args_in)
  }
fn_run_tool = function(step, tool, args, run_label){
  cat("\n---", run_label, "---\n", path_tools[[tool]], paste(args, collapse = " "), "\n")
  result = try(system2(path_tools[[tool]], args = args, stdout = TRUE, stderr = TRUE), silent = TRUE)
  exit_status = attr(result, "status")
  exit_status = if(is.null(exit_status)) 0L else exit_status
  if(inherits(result, "try-error")) exit_status = -1L
  cat(paste(result, collapse = "\n"), "\n")
  fn_check(step, tool, paste0(run_label, ": exits with status 0"), exit_status == 0, exit_status, 0)
  env_checks$commands[[length(env_checks$commands) + 1]] = data.table::data.table(
      step        = step
    , tool        = tool
    , run_label   = run_label
    , command     = paste(c(tool, fn_portable_args(args)), collapse = " ")
    , exit_status = exit_status
    )
  invisible(list(exit_status = exit_status, console = as.character(result)))
  }

#4. define a helper that checks an output file exists and isn't empty
fn_check_file = function(step, tool, path, what){
  exists_nonempty = file.exists(path) && file.info(path)$size > 0
  fn_check(step, tool, paste0(what, " written"), exists_nonempty, if(exists_nonempty) basename(path) else "missing", basename(path))
  exists_nonempty
  }

#5. define a helper that checks a raster covers the tile's real
#   coordinates
#  - a grid may run past the tile edge by up to one cell when the tile
#    width isn't a multiple of the cell size; anything beyond that means
#    the raster is misplaced
fn_check_raster_extent = function(step, tool, r, what){
  ext_r = as.vector(terra::ext(r))
  cell_max = max(terra::res(r))
  off_left   = ext_tile[["xmin"]] - ext_r[["xmin"]]
  off_right  = ext_r[["xmax"]] - ext_tile[["xmax"]]
  off_bottom = ext_tile[["ymin"]] - ext_r[["ymin"]]
  off_top    = ext_r[["ymax"]] - ext_tile[["ymax"]]
  offsets = c(off_left, off_right, off_bottom, off_top)
  within_one_cell = all(offsets > -0.01 & offsets <= cell_max + 0.01)
  fn_check(step, tool, paste0(what, ": raster sits on the tile's real coordinates"), within_one_cell, round(ext_r, 2), round(ext_tile, 2))
  }

#6. define a helper that checks a raster has a coordinate system attached
fn_check_raster_crs = function(step, tool, r, what){
  has_crs = nzchar(terra::crs(r))
  fn_check(step, tool, paste0(what, ": coordinate system attached"), has_crs, if(has_crs) "present" else "none", "present", severity = "soft")
  }

#7. define a helper that summarizes every band of a raster in one table
fn_raster_overview = function(r){
  v = terra::values(r)
  data.table::data.table(
      band    = names(r)
    , n_cells = nrow(v)
    , n_na    = colSums(is.na(v))
    , min     = suppressWarnings(apply(v, 2, min, na.rm = TRUE))
    , mean    = suppressWarnings(apply(v, 2, mean, na.rm = TRUE))
    , max     = suppressWarnings(apply(v, 2, max, na.rm = TRUE))
    )
  }

#8. define a helper that reads a LAS/LAZ output quietly, or returns NULL
fn_read_las = function(path){
  if(!file.exists(path)) return(NULL)
  suppressWarnings(lidR::readLAS(path))
  }

#9. define a helper that builds a reference grid on the tile's own
#   coordinates, for rasterizing points in R the same way the tools should
#  - origin at the tile's lower-left corner, rows counted down from the
#    top, one extra row/column when the width isn't a multiple of the cell
fn_reference_grid = function(cellsize){
  n_col = ceiling((ext_tile[["xmax"]] - ext_tile[["xmin"]]) / cellsize)
  n_row = ceiling((ext_tile[["ymax"]] - ext_tile[["ymin"]]) / cellsize)
  terra::rast(
      xmin = ext_tile[["xmin"]]
    , xmax = ext_tile[["xmin"]] + n_col * cellsize
    , ymin = ext_tile[["ymax"]] - n_row * cellsize
    , ymax = ext_tile[["ymax"]]
    , resolution = cellsize
    , crs = crs_tile_wkt
    )
  }

#10. define a helper that looks up a ground raster's height under each
#    point, returning NA for points outside the raster
fn_ground_at_points = function(r_ground, x, y){
  terra::extract(r_ground, cbind(x, y), method = "bilinear")[ , 1]
  }

#11. define a helper that plots a raster band as a map, with an optional
#    set of highlighted cells
fn_plot_raster_map = function(r, band, title, highlight_above = NULL){
  df_r = as.data.frame(r[[band]], xy = TRUE, na.rm = FALSE)
  names(df_r) = c("x", "y", "value")
  p = ggplot2::ggplot(df_r, ggplot2::aes(x = x, y = y, fill = value)) +
    ggplot2::geom_raster() +
    ggplot2::scale_fill_viridis_c(name = band, na.value = "grey85") +
    ggplot2::coord_equal() +
    ggplot2::labs(title = title, x = "x", y = "y") +
    ggplot2::theme_minimal(base_size = 11)
  if(!is.null(highlight_above)) {
    df_hi = df_r[!is.na(df_r$value) & df_r$value > highlight_above, ]
    p = p + ggplot2::geom_tile(data = df_hi, ggplot2::aes(x = x, y = y), fill = NA, colour = "red", linewidth = 0.4, inherit.aes = FALSE)
    }
  p
  }
```

# Overview

This script runs each of the 15 FUSION2 tools on the example tile. It then
checks every output against an independent result computed in R from the
same points. Each check asks one of three questions:

1. Was the output written, and is it on the tile's real coordinates?
2. Do counts match exactly what the options asked for? For example, a
   first-returns filter should keep exactly the tile's first returns.
3. Are the values physically plausible? For example, no height above
   ground should exceed the tallest point in the tile.

Step 1 reads the tile in R and builds the reference values the later
steps compare against. Steps 2-16 run one tool each, in the order a real
workflow would call them: point-cloud tools first, then the ground DEM,
then the tools that depend on it. Step 17 gathers every check into one
table.

Two groups of tool options are covered by the earlier scripts rather than
repeated here: `gridmetrics`' batch/tiled mode and SQLite table output
(001 and 003), and `pipeline`'s many stage chains (003 and 004).

# Step 1: Read the tile in R and build reference values

The references below come from `lidR` and `terra` rather than FUSION2, so
agreement means two independent implementations reached the same answer.
The reference ground surface uses `lidR`'s triangulation of the tile's own
ground-classified points (class 2). Heights above ground are computed
against that surface.

```{r}
#| label: step-1-reference

#1. read the tile and record its basic properties
#  - ASPRS LAS standard specifies that points with the withheld flag set
#    should be treated as deleted during processing. FUSION2 omits
#    withheld points when reading point clouds. We retain both the raw
#    total count (for file-header tools like catalog) and the non-withheld
#    points for processing tools.
las_raw = suppressWarnings(lidR::readLAS(path_las))
las = suppressWarnings(lidR::filter_poi(las_raw, !Withheld_flag))
n_points_raw = lidR::npoints(las_raw)
n_points     = lidR::npoints(las)
ext_tile = c(
    xmin = las_raw@header$`Min X`
  , xmax = las_raw@header$`Max X`
  , ymin = las_raw@header$`Min Y`
  , ymax = las_raw@header$`Max Y`
  )
z_range_raw = c(las_raw@header$`Min Z`, las_raw@header$`Max Z`)
z_range     = range(las@data$Z)
crs_tile_wkt = sf::st_as_text(sf::st_crs(las_raw))

#2. tabulate returns, classes, and withheld points
#  - "withheld" points are flagged in the LAS file as ones that should be
#    treated as deleted; later steps check whether the tools drop them
dt_tile_summary = data.table::data.table(
    n_points_raw   = n_points_raw
  , n_points_valid = n_points
  , n_first_return = sum(las$ReturnNumber == 1)
  , n_ground       = sum(las$Classification == 2)
  , n_noise        = sum(las_raw$Classification == 7)
  , n_withheld     = sum(las_raw$Withheld_flag)
  , z_min          = z_range[1]
  , z_max          = z_range[2]
  )
print(ext_tile)
print(dt_tile_summary)

#3. build the reference ground surface and heights above ground
#  - the tallest reference height sets the upper limit for any height a
#    tool reports; 5 ft of slack allows for interpolation differences
#    between lidR's triangulation and FUSION2's ground model
las_ground = lidR::filter_poi(las, Classification == 2L)
r_ground_ref = suppressWarnings(lidR::rasterize_terrain(las, res = 5, algorithm = lidR::tin()))
las_norm = suppressWarnings(lidR::normalize_height(las, algorithm = lidR::tin()))
height_max_ref = max(las_norm$Z)
height_limit = height_max_ref + 5
cat("Tallest reference height above ground:", round(height_max_ref, 2), "ft; plausibility limit:", round(height_limit, 2), "ft\n")

#4. interrogate the references before relying on them
#  - the ground surface must span the tile's elevation range, and no
#    point should sit far below it
if(any(is.na(las_norm$Z))) stop("Reference height normalization left NA heights.")
if(min(las_norm$Z) < -5) warning("Some points sit more than 5 ft below the reference ground surface.")
fn_check(1, "reference", "reference heights stay within a plausible range", height_max_ref < 300 & min(las_norm$Z) > -10, round(range(las_norm$Z), 2), "-10 to 300 ft")
```

# Step 2: Summarize the tile with `catalog`

`catalog` reports each input file's point count and bounding box, and with
`/density:<cellsize>` also writes a point-density raster.

```{r}
#| label: step-2-catalog

#1. run catalog with both the CSV report and the density raster requested
dir_catalog = file.path(dir_run, "catalog")
dir.create(dir_catalog, showWarnings = FALSE)
path_catalog_csv = file.path(dir_catalog, "catalog.csv")
fn_run_tool(2, "catalog", c(path_las, paste0("/output:", path_catalog_csv), "/density:10"), "catalog")

#2. check the CSV report against the tile read in R
#  - bounds should match the LAS header to the header's own precision
#    (0.01 ft); a larger gap means the CSV rounded the coordinates
if(fn_check_file(2, "catalog", path_catalog_csv, "CSV report")) {
  dt_catalog = data.table::fread(path_catalog_csv)
  print(dt_catalog)
  fn_check(2, "catalog", "point count matches the tile", dt_catalog$PointCount == n_points_raw, dt_catalog$PointCount, n_points_raw)
  bounds_catalog = c(dt_catalog$MinX, dt_catalog$MaxX, dt_catalog$MinY, dt_catalog$MaxY)
  bounds_tile    = c(ext_tile[["xmin"]], ext_tile[["xmax"]], ext_tile[["ymin"]], ext_tile[["ymax"]])
  fn_check(2, "catalog", "XY bounds match the tile within 0.01 ft", all(abs(bounds_catalog - bounds_tile) <= 0.01), bounds_catalog, bounds_tile)
  fn_check(2, "catalog", "Z range matches the tile within 0.01 ft", all(abs(c(dt_catalog$MinZ, dt_catalog$MaxZ) - z_range_raw) <= 0.01), c(dt_catalog$MinZ, dt_catalog$MaxZ), z_range_raw)
  }

#3. check that /density produced a raster placed on the tile, with a
#   coordinate system, whose cell counts sum exactly to the point count
#  - a raster with the right counts can still sit at the grid origin
#    (0, 0) instead of on the tile, so placement is checked separately
#  - points exactly on the tile's east or south edge belong to the last
#    column or row, so every point is counted and the sum is exact
paths_catalog_tif = list.files(dir_catalog, pattern = "\\.tif$", full.names = TRUE)
fn_check(2, "catalog", "/density raster written", length(paths_catalog_tif) > 0, length(paths_catalog_tif), ">= 1 .tif")
if(length(paths_catalog_tif) > 0) {
  r_catalog_density = terra::rast(paths_catalog_tif[1])
  fn_check_raster_extent(2, "catalog", r_catalog_density, "density raster")
  fn_check_raster_crs(2, "catalog", r_catalog_density, "density raster")
  n_implied = sum(terra::values(r_catalog_density[[1]]), na.rm = TRUE)
  fn_check(2, "catalog", "density raster sums to the point count", n_implied == n_points_raw, round(n_implied), n_points_raw)
  }
```

# Step 3: Filter points with `filterdata`

`filterdata` keeps points by return number, classification, or elevation
band. Each run below is checked two ways: the kept count must equal the
count R finds with the same rule, and every kept point must satisfy the
rule.

```{r}
#| label: step-3-filterdata

#1. define the three filters and the rule R uses to reproduce each
#  - the elevation band is the 20 ft band centred on the tile's median
#    point elevation, so it always keeps a sizeable share of the points
dir_filter = file.path(dir_run, "filterdata")
dir.create(dir_filter, showWarnings = FALSE)
z_band = round(stats::median(las$Z)) + c(-10, 10)
dt_filters = data.table::data.table(
    run  = c("first_returns", "ground_class", "z_band")
  , args = list("/return:1", "/class:2", c(paste0("/minz:", z_band[1]), paste0("/maxz:", z_band[2])))
  )
fn_filter_rule = function(run, x){
  if(run == "first_returns") return(x$ReturnNumber == 1)
  if(run == "ground_class")  return(x$Classification == 2)
  x$Z >= z_band[1] & x$Z <= z_band[2]
  }

#2. run each filter and check it
for(i in seq_len(nrow(dt_filters))) {
  run_i = dt_filters$run[i]
  path_out_i = file.path(dir_filter, paste0(run_i, ".las"))
  fn_run_tool(3, "filterdata", c(path_las, paste0("/output:", path_out_i), dt_filters$args[[i]]), paste0("filterdata ", run_i))

  #  - count kept vs. count R expects
  n_expected_i = sum(fn_filter_rule(run_i, las))
  las_out_i = fn_read_las(path_out_i)
  if(fn_check(3, "filterdata", paste0(run_i, ": output readable"), !is.null(las_out_i), if(is.null(las_out_i)) "unreadable" else "ok", "ok")) {
    fn_check(3, "filterdata", paste0(run_i, ": kept count matches R"), lidR::npoints(las_out_i) == n_expected_i, lidR::npoints(las_out_i), n_expected_i)
    fn_check(3, "filterdata", paste0(run_i, ": every kept point satisfies the filter"), all(fn_filter_rule(run_i, las_out_i)), sum(!fn_filter_rule(run_i, las_out_i)), "0 violations")
    fn_check(3, "filterdata", paste0(run_i, ": coordinate system kept"), !is.na(sf::st_crs(las_out_i)), if(is.na(sf::st_crs(las_out_i))) "none" else "present", "present")
    fn_check(3, "filterdata", paste0(run_i, ": withheld points dropped"), sum(las_out_i$Withheld_flag) == 0, sum(las_out_i$Withheld_flag), 0, severity = "soft")
    }
  }
```

# Step 4: Thin points with `thindata`

`thindata` keeps at most one point per grid cell. Its grid lines fall on
whole multiples of the cell size (x = 0, 2, 4, ... ft), not on the tile's
own corner. The check rebuilds that same 2 ft grid in R and confirms that
no cell holds two kept points and that the number kept equals the number
of occupied cells. It also confirms that every kept point exists in the
original tile.

```{r}
#| label: step-4-thindata

#1. run thindata at a 2 ft cell size
cellsize_thin = 2
path_thin = file.path(dir_run, "thindata.las")
fn_run_tool(4, "thindata", c(path_las, paste0("/output:", path_thin), paste0("/cellsize:", cellsize_thin)), "thindata")

#2. assign every point (original and kept) to its 2 ft cell, on a grid
#   whose lines fall on whole multiples of the cell size
fn_cell_id = function(x, y){
  paste(floor(x / cellsize_thin), floor(y / cellsize_thin))
  }
n_cells_occupied = data.table::uniqueN(fn_cell_id(las$X, las$Y))

#3. check the kept points
las_thin = fn_read_las(path_thin)
if(fn_check(4, "thindata", "output readable", !is.null(las_thin), if(is.null(las_thin)) "unreadable" else "ok", "ok")) {
  cells_thin = fn_cell_id(las_thin$X, las_thin$Y)
  fn_check(4, "thindata", "no cell keeps more than one point", !anyDuplicated(cells_thin), sum(duplicated(cells_thin)), "0 duplicate cells")
  fn_check(4, "thindata", "kept count equals occupied cells", lidR::npoints(las_thin) == n_cells_occupied, lidR::npoints(las_thin), n_cells_occupied)
  key_tile = paste(las$X, las$Y, las$Z)
  key_thin = paste(las_thin$X, las_thin$Y, las_thin$Z)
  fn_check(4, "thindata", "every kept point exists in the tile", all(key_thin %in% key_tile), sum(!key_thin %in% key_tile), "0 new points")
  }
```

# Step 5: Build a ground DEM with `groundfilter`

`groundfilter` interpolates a ground surface raster, and with
`/output-points` also writes the points it classed as ground. Every later
tool that reports height above ground uses this DEM, so the script stops
here if the DEM is missing.

```{r}
#| label: step-5-groundfilter

#1. run groundfilter with both outputs requested
cellsize_ground = 5
path_ground_dem = file.path(dir_run, "ground_dem.tif")
path_ground_pts = file.path(dir_run, "ground_points.las")
fn_run_tool(5, "groundfilter", c(
    path_las
  , paste0("/output-raster:", path_ground_dem)
  , paste0("/output-points:", path_ground_pts)
  , paste0("/cellsize:", cellsize_ground)
  ), "groundfilter")

#2. stop if the DEM is missing -- later steps can't run without it
if(!fn_check_file(5, "groundfilter", path_ground_dem, "ground DEM")) stop("groundfilter wrote no ground DEM; later steps depend on it.")
r_ground = terra::rast(path_ground_dem)
print(fn_raster_overview(r_ground))

#3. check the DEM's placement, completeness, and values
fn_check_raster_extent(5, "groundfilter", r_ground, "ground DEM")
fn_check_raster_crs(5, "groundfilter", r_ground, "ground DEM")
v_ground = terra::values(r_ground)[ , 1]
fn_check(5, "groundfilter", "ground DEM has no empty cells", !any(is.na(v_ground)), sum(is.na(v_ground)), 0)
z_ground_pts = range(las_ground$Z)
fn_check(5, "groundfilter", "ground DEM stays within the ground points' elevation range (+/- 1 ft)", min(v_ground, na.rm = TRUE) >= z_ground_pts[1] - 1 & max(v_ground, na.rm = TRUE) <= z_ground_pts[2] + 1, round(range(v_ground, na.rm = TRUE), 2), round(z_ground_pts, 2))

#4. compare against the lidR reference ground surface, cell by cell
#  - the reference is resampled onto the FUSION2 grid, so the comparison
#    only covers cells both rasters cover
r_ground_ref_aligned = terra::resample(r_ground_ref, r_ground, method = "bilinear")
diff_ground = abs(terra::values(r_ground)[ , 1] - terra::values(r_ground_ref_aligned)[ , 1])
fn_check(5, "groundfilter", "median |FUSION2 - lidR| ground difference under 1 ft", stats::median(diff_ground, na.rm = TRUE) < 1, round(stats::median(diff_ground, na.rm = TRUE), 3), "< 1 ft")
fn_check(5, "groundfilter", "95th percentile |FUSION2 - lidR| ground difference under 3 ft", stats::quantile(diff_ground, 0.95, na.rm = TRUE) < 3, round(stats::quantile(diff_ground, 0.95, na.rm = TRUE), 3), "< 3 ft")

#5. check the ground-points output
las_ground_out = fn_read_las(path_ground_pts)
if(fn_check(5, "groundfilter", "/output-points LAS written", !is.null(las_ground_out), if(is.null(las_ground_out)) "missing" else "present", "present")) {
  fn_check(5, "groundfilter", "/output-points holds only ground-classed points", all(las_ground_out$Classification == 2), paste(sort(unique(las_ground_out$Classification)), collapse = ","), "2")
  }

#6. plot FUSION2 ground (x) against the lidR reference (y)
dt_ground_cmp = data.table::data.table(fusion2 = terra::values(r_ground)[ , 1], lidr_reference = terra::values(r_ground_ref_aligned)[ , 1])
ggplot2::ggplot(dt_ground_cmp, ggplot2::aes(x = fusion2, y = lidr_reference)) +
  ggplot2::geom_abline(slope = 1, intercept = 0, colour = "grey50") +
  ggplot2::geom_point(alpha = 0.5) +
  ggplot2::coord_equal() +
  ggplot2::labs(title = "Ground DEM: FUSION2 vs. lidR reference", x = "FUSION2 groundfilter (ft)", y = "lidR TIN reference (ft)") +
  ggplot2::theme_minimal(base_size = 11)
```

# Step 6: Clip points with `clipdata`

`clipdata` keeps points inside a bounding box, inside polygons, or outside
polygons. With `/ground`, the `/minz`/`/maxz` limits apply to height above
ground rather than raw elevation. Two small circular plots are drawn in R
to exercise the polygon options.

```{r}
#| label: step-6-clipdata

dir_clip = file.path(dir_run, "clipdata")
dir.create(dir_clip, showWarnings = FALSE)

#1. clip to a 100 x 100 ft box and check it against R
#  - the box starts 25 ft in from the tile's lower-left corner
box_clip = c(
    xmin = ext_tile[["xmin"]] + 25
  , ymin = ext_tile[["ymin"]] + 25
  , xmax = ext_tile[["xmin"]] + 125
  , ymax = ext_tile[["ymin"]] + 125
  )
arg_box = paste0("/extent:", paste(box_clip[c("xmin", "ymin", "xmax", "ymax")], collapse = ","))
path_clip_box = file.path(dir_clip, "box.las")
fn_run_tool(6, "clipdata", c(path_las, paste0("/output:", path_clip_box), arg_box), "clipdata box")
in_box = las$X >= box_clip[["xmin"]] & las$X <= box_clip[["xmax"]] & las$Y >= box_clip[["ymin"]] & las$Y <= box_clip[["ymax"]]
las_clip_box = fn_read_las(path_clip_box)
if(fn_check(6, "clipdata", "box: output readable", !is.null(las_clip_box), if(is.null(las_clip_box)) "unreadable" else "ok", "ok")) {
  fn_check(6, "clipdata", "box: kept count matches R", lidR::npoints(las_clip_box) == sum(in_box), lidR::npoints(las_clip_box), sum(in_box))
  }

#2. clip the same box by height above ground (2-200 ft) and check it
#  - heights are recomputed in R against the FUSION2 ground DEM itself,
#    so this checks the clip logic, not the DEM
path_clip_ht = file.path(dir_clip, "box_height_2_200.las")
fn_run_tool(6, "clipdata", c(path_las, paste0("/output:", path_clip_ht), arg_box, paste0("/ground:", path_ground_dem), "/minz:2", "/maxz:200"), "clipdata box + height")
height_vs_fusion_ground = las$Z - fn_ground_at_points(r_ground, las$X, las$Y)
n_expected_ht = sum(in_box & height_vs_fusion_ground >= 2 & height_vs_fusion_ground <= 200, na.rm = TRUE)
las_clip_ht = fn_read_las(path_clip_ht)
if(fn_check(6, "clipdata", "box + height: output readable", !is.null(las_clip_ht), if(is.null(las_clip_ht)) "unreadable" else "ok", "ok")) {
  height_out = las_clip_ht$Z - fn_ground_at_points(r_ground, las_clip_ht$X, las_clip_ht$Y)
  fn_check(6, "clipdata", "box + height: kept count within 10% of R", abs(lidR::npoints(las_clip_ht) - n_expected_ht) <= max(2, 0.10 * n_expected_ht), lidR::npoints(las_clip_ht), n_expected_ht)
  fn_check(6, "clipdata", "box + height: every kept point is 2-200 ft above ground (+/- 0.5 ft)", all(height_out >= 1.5 & height_out <= 200.5, na.rm = TRUE), round(range(height_out, na.rm = TRUE), 2), "1.5 to 200.5 ft")
  }

#3. build two 30 ft radius circular plots as a shapefile
#  - centred 50 ft and 150 ft in from the tile's lower-left corner, so
#    both circles sit wholly inside the tile
path_plots_shp = file.path(dir_clip, "example_plots.shp")
sf_plots = sf::st_sf(
    plot_id = c("plotA", "plotB")
  , geometry = sf::st_sfc(
        sf::st_buffer(sf::st_point(c(ext_tile[["xmin"]] + 50, ext_tile[["ymin"]] + 50)), 30)
      , sf::st_buffer(sf::st_point(c(ext_tile[["xmin"]] + 150, ext_tile[["ymin"]] + 150)), 30)
      , crs = sf::st_crs(las)
      )
  )
sf::st_write(sf_plots, path_plots_shp, quiet = TRUE, append = FALSE)

#4. count points inside each plot in R
sf_pts = sf::st_as_sf(data.table::data.table(X = las$X, Y = las$Y), coords = c("X", "Y"), crs = sf::st_crs(las))
mat_in_plot = sf::st_within(sf_pts, sf_plots, sparse = FALSE)
n_in_plot = colSums(mat_in_plot)
names(n_in_plot) = sf_plots$plot_id

#5. clip one file per plot, named by plot_id, and check each count
#  - points exactly on a circle's edge can fall either way, so counts may
#    differ by a handful of points
dir_clip_multi = file.path(dir_clip, "per_plot")
dir.create(dir_clip_multi, showWarnings = FALSE)
fn_run_tool(6, "clipdata", c(path_las, paste0("/output:", dir_clip_multi), paste0("/shape:", path_plots_shp), "/multifile", "/field:plot_id"), "clipdata /multifile")
for(plot_i in names(n_in_plot)) {
  path_plot_i = list.files(dir_clip_multi, pattern = paste0("^", plot_i, "\\.la[sz]$"), full.names = TRUE)
  fn_check(6, "clipdata", paste0("/multifile: ", plot_i, " file named from /field"), length(path_plot_i) == 1, length(path_plot_i), 1)
  if(length(path_plot_i) == 1) {
    n_plot_i = lidR::npoints(fn_read_las(path_plot_i))
    fn_check(6, "clipdata", paste0("/multifile: ", plot_i, " count within 0.5% of R"), abs(n_plot_i - n_in_plot[[plot_i]]) <= max(3, 0.005 * n_in_plot[[plot_i]]), n_plot_i, n_in_plot[[plot_i]])
    }
  }

#6. clip everything outside both plots and check it is the complement
path_clip_outside = file.path(dir_clip, "outside_plots.las")
fn_run_tool(6, "clipdata", c(path_las, paste0("/output:", path_clip_outside), paste0("/shape:", path_plots_shp), "/outside"), "clipdata /outside")
las_clip_outside = fn_read_las(path_clip_outside)
if(fn_check(6, "clipdata", "/outside: output readable", !is.null(las_clip_outside), if(is.null(las_clip_outside)) "unreadable" else "ok", "ok")) {
  n_outside_expected = n_points - sum(rowSums(mat_in_plot) > 0)
  fn_check(6, "clipdata", "/outside: count equals tile minus in-plot points (within 0.5%)", abs(lidR::npoints(las_clip_outside) - n_outside_expected) <= 0.005 * n_outside_expected, lidR::npoints(las_clip_outside), n_outside_expected)
  }
```

# Step 7: Build canopy height models with `canopymodel`

`canopymodel` grids the highest point in each cell and subtracts the
ground DEM. Three versions are built: plain, smoothed (`/smooth:3`), and
slope-adjusted (`/slope`). The reference canopy model rasterizes the same
highest points in R, measured against the FUSION2 ground DEM, onto the
same grid.

The most important check is the height ceiling. No canopy cell may be
taller than the tallest point above ground in the tile. A cell far above
that ceiling holds a raw elevation, meaning the ground was never
subtracted.

```{r}
#| label: step-7-canopymodel

#1. build the three canopy height models
cellsize_chm = 3
path_chm = file.path(dir_run, "chm.tif")
path_chm_smooth = file.path(dir_run, "chm_smooth3.tif")
path_chm_slope = file.path(dir_run, "chm_slope.tif")
args_chm_common = c(path_las, paste0("/ground:", path_ground_dem), paste0("/cellsize:", cellsize_chm))
fn_run_tool(7, "canopymodel", c(args_chm_common, paste0("/output:", path_chm)), "canopymodel")
fn_run_tool(7, "canopymodel", c(args_chm_common, paste0("/output:", path_chm_smooth), "/smooth:3"), "canopymodel /smooth:3")
fn_run_tool(7, "canopymodel", c(args_chm_common, paste0("/output:", path_chm_slope), "/slope"), "canopymodel /slope")

#2. stop if the plain model is missing -- Steps 13-15 depend on it
if(!fn_check_file(7, "canopymodel", path_chm, "canopy height model")) stop("canopymodel wrote no CHM; later steps depend on it.")
r_chm = terra::rast(path_chm)
print(fn_raster_overview(r_chm))
fn_check_raster_extent(7, "canopymodel", r_chm, "CHM")
fn_check_raster_crs(7, "canopymodel", r_chm, "CHM")

#3. check the height range
#  - small negative heights are normal (a return slightly below the
#    interpolated ground); more than 1 ft below is not
v_chm = terra::values(r_chm)[ , 1]
n_chm_too_tall = sum(v_chm > height_limit, na.rm = TRUE)
fn_check(7, "canopymodel", "no CHM cell exceeds the tallest point above ground", n_chm_too_tall == 0, paste0(n_chm_too_tall, " cells; max ", round(max(v_chm, na.rm = TRUE), 2)), paste0("0 cells above ", round(height_limit, 2), " ft"))
fn_check(7, "canopymodel", "no CHM cell more than 1 ft below ground", min(v_chm, na.rm = TRUE) >= -1, round(min(v_chm, na.rm = TRUE), 2), ">= -1 ft")

#4. compare with the reference canopy model on the same grid
#  - the comparison is limited to cells both models consider plausible,
#    so the height-ceiling failures above don't mask agreement elsewhere
height_above_fusion_ground = las$Z - fn_ground_at_points(r_ground, las$X, las$Y)
v_pts = terra::vect(cbind(las$X, las$Y), crs = crs_tile_wkt, atts = data.frame(h = height_above_fusion_ground))
r_chm_ref = terra::rasterize(v_pts, r_chm, field = "h", fun = max)
v_chm_ref = terra::values(r_chm_ref)[ , 1]
keep_cmp = !is.na(v_chm) & !is.na(v_chm_ref) & v_chm <= height_limit
diff_chm = abs(v_chm[keep_cmp] - v_chm_ref[keep_cmp])
fn_check(7, "canopymodel", "median |CHM - R reference| under 0.5 ft (plausible cells)", stats::median(diff_chm) < 0.5, round(stats::median(diff_chm), 3), "< 0.5 ft")

#5. check the smoothed and slope-adjusted variants actually changed things
#  - smoothing should lower the cell-to-cell variation; /slope should
#    change at least some cells
if(fn_check_file(7, "canopymodel", path_chm_smooth, "/smooth:3 CHM")) {
  v_smooth = terra::values(terra::rast(path_chm_smooth))[ , 1]
  fn_check(7, "canopymodel", "/smooth:3 lowers cell-to-cell variance", stats::var(v_smooth, na.rm = TRUE) < stats::var(v_chm, na.rm = TRUE), round(stats::var(v_smooth, na.rm = TRUE), 2), paste0("< ", round(stats::var(v_chm, na.rm = TRUE), 2)))
  }
if(fn_check_file(7, "canopymodel", path_chm_slope, "/slope CHM")) {
  v_slope = terra::values(terra::rast(path_chm_slope))[ , 1]
  fn_check(7, "canopymodel", "/slope changes at least one cell", max(abs(v_slope - v_chm), na.rm = TRUE) > 0, round(max(abs(v_slope - v_chm), na.rm = TRUE), 3), "> 0")
  }

#6. map the CHM, outlining any cells above the height ceiling in red
fn_plot_raster_map(r_chm, 1, "Canopy height model (red = above the height ceiling)", highlight_above = height_limit)
```

# Step 8: Map point density with `returndensity`

`returndensity` writes point density per unit area and the percent of
returns that are first returns. Multiplying density by cell area and
summing should give back the tile's point count exactly.

```{r}
#| label: step-8-returndensity

#1. run returndensity at a 10 ft cell size
path_returndensity = file.path(dir_run, "returndensity.tif")
fn_run_tool(8, "returndensity", c(path_las, paste0("/output:", path_returndensity), "/cellsize:10"), "returndensity")

if(fn_check_file(8, "returndensity", path_returndensity, "density raster")) {
  #2. overview and placement
  r_rd = terra::rast(path_returndensity)
  print(fn_raster_overview(r_rd))
  fn_check_raster_extent(8, "returndensity", r_rd, "density raster")
  fn_check_raster_crs(8, "returndensity", r_rd, "density raster")

  #3. check density x cell area against the point count, overall and per cell
  cell_area = prod(terra::res(r_rd))
  v_density = terra::values(r_rd[[1]])[ , 1]
  n_implied = sum(v_density, na.rm = TRUE) * cell_area
  fn_check(8, "returndensity", "density x cell area sums to the point count (within 0.5%)", abs(n_implied - n_points) / n_points < 0.005, round(n_implied), n_points)
  r_count_ref = terra::rasterize(v_pts, r_rd, fun = "length")
  v_count_ref = terra::values(r_count_ref)[ , 1]
  v_count_ref[is.na(v_count_ref)] = 0
  fn_check(8, "returndensity", "per-cell counts match R exactly", all(abs(v_density * cell_area - v_count_ref) < 0.5), sum(abs(v_density * cell_area - v_count_ref) >= 0.5), "0 mismatched cells")

  #4. check the first-return percentage and the band's unit label
  #  - the tile's units are US survey feet, so a band labelled per square
  #    meter is mislabelled unless the tool converted units
  v_first_pct = terra::values(r_rd[[2]])[ , 1]
  fn_check(8, "returndensity", "first-return percent within 0-100", all(v_first_pct >= 0 & v_first_pct <= 100, na.rm = TRUE), round(range(v_first_pct, na.rm = TRUE), 2), "0 to 100")
  fn_check(8, "returndensity", "density band's unit label matches the tile's linear units (feet)", !grepl("m2", names(r_rd)[1]), names(r_rd)[1], "no m2 label on a feet-based tile", severity = "soft")
  }
```

# Step 9: Map density by height stratum with `densitymetrics`

`densitymetrics` counts returns in height bands above ground (default
breaks 0.5, 2, 5, 10, and 20 ft, giving six bands) and writes one raster
band per stratum. Per cell, the stratum counts should add up to that
cell's total returns. On this forested tile the top stratum (above 20 ft)
holds most of the canopy returns, so its count against R is a meaningful
check rather than a comparison of two empty counts.

```{r}
#| label: step-9-densitymetrics

#1. run densitymetrics with the FUSION2 ground DEM
dir_dm = file.path(dir_run, "densitymetrics")
dir.create(dir_dm, showWarnings = FALSE)
#  - /output-table asks for the per-cell CSV that step 3 below checks;
#    densitymetrics writes no table unless it is given
path_dm_csv = file.path(dir_dm, "dm_densitymetrics.csv")
fn_run_tool(9, "densitymetrics", c(path_las, path_ground_dem, paste0("/outdir:", dir_dm), "/output:dm", "/cellsize:10", paste0("/output-table:", path_dm_csv)), "densitymetrics")

#2. overview and placement of the raster
path_dm_tif = list.files(dir_dm, pattern = "\\.tif$", full.names = TRUE)[1]
if(fn_check_file(9, "densitymetrics", if(is.na(path_dm_tif)) file.path(dir_dm, "dm_densitymetrics.tif") else path_dm_tif, "stratum raster")) {
  r_dm = terra::rast(path_dm_tif)
  print(fn_raster_overview(r_dm))
  fn_check_raster_extent(9, "densitymetrics", r_dm, "stratum raster")
  fn_check_raster_crs(9, "densitymetrics", r_dm, "stratum raster")
  }

#3. check the per-cell CSV table
#  - the table was requested with /output-table in step 1, so a missing
#    file is a hard failure, and the content checks below must all run
if(fn_check(9, "densitymetrics", "per-cell CSV table written", file.exists(path_dm_csv), if(file.exists(path_dm_csv)) "present" else "none", "present")) {
  dt_dm = data.table::fread(path_dm_csv)
  nms_strata_cnt = grep("^StrataCnt_", names(dt_dm), value = TRUE)
  stratum_sum = rowSums(dt_dm[ , ..nms_strata_cnt])
  fn_check(9, "densitymetrics", "per cell, stratum counts add to total returns", all(stratum_sum == dt_dm$TotalReturns), sum(stratum_sum != dt_dm$TotalReturns), "0 mismatched cells")
  fn_check(9, "densitymetrics", "total returns across cells equals the point count", sum(dt_dm$TotalReturns) == n_points, sum(dt_dm$TotalReturns), n_points)

  #  - the top stratum's count against R's count of points more than 20 ft
  #    above the FUSION2 ground DEM
  n_top_expected = sum(height_above_fusion_ground > 20, na.rm = TRUE)
  n_top_observed = sum(dt_dm[[nms_strata_cnt[length(nms_strata_cnt)]]])
  fn_check(9, "densitymetrics", "top stratum (> 20 ft) count matches R (within 5 points)", abs(n_top_observed - n_top_expected) <= 5, n_top_observed, n_top_expected)
  }
```

# Step 10: Compute the full metric grid with `gridmetrics`

`gridmetrics` computes the full elevation and intensity statistic bundle
per cell, plus height strata (`/strata`, `/strataraster`) and surface
roughness (`/surfstats`). Script 001 covers the table-export options in
depth. The checks here focus on placement, per-cell return counts, the
height ceiling, and agreement between the raster and its CSV table.

```{r}
#| label: step-10-gridmetrics

#1. run gridmetrics with strata, strata rasters, and surface stats
dir_gm = file.path(dir_run, "gridmetrics")
dir.create(dir_gm, showWarnings = FALSE)
path_gm_table = file.path(dir_gm, "gridmetrics_table.csv")
fn_run_tool(10, "gridmetrics", c(
    path_las
  , path_ground_dem
  , "/cellsize:10"
  , paste0("/outdir:", dir_gm)
  , "/strata:0.5,2,5,10,20"
  , "/strataraster"
  , "/surfstats"
  , paste0("/output-table:", path_gm_table)
  ), "gridmetrics")

#2. find the raster and table
#  - older builds ignore /output-table and write their own CSV name, so
#    either name is accepted here; the soft check records which happened
path_gm_tif = list.files(dir_gm, pattern = "_gridmetrics\\.tif$", full.names = TRUE)[1]
path_gm_csv = if(file.exists(path_gm_table)) path_gm_table else list.files(dir_gm, pattern = "\\.csv$", full.names = TRUE)[1]
fn_check(10, "gridmetrics", "/output-table path honored", file.exists(path_gm_table), if(file.exists(path_gm_table)) "written" else "ignored", "written", severity = "soft")

if(fn_check_file(10, "gridmetrics", if(is.na(path_gm_tif)) file.path(dir_gm, "missing_gridmetrics.tif") else path_gm_tif, "metrics raster")) {
  #3. overview and placement
  r_gm = terra::rast(path_gm_tif)
  dt_gm_overview = fn_raster_overview(r_gm)
  print(dt_gm_overview[ , .(n_bands = .N, n_all_na = sum(n_na == n_cells))])
  fn_check_raster_extent(10, "gridmetrics", r_gm, "metrics raster")
  fn_check_raster_crs(10, "gridmetrics", r_gm, "metrics raster")

  #4. list any band that is empty everywhere
  #  - on this closed-canopy tile every band, including the surface
  #    roughness bands, should hold values somewhere
  nms_all_na = dt_gm_overview[n_na == n_cells, band]
  fn_check(10, "gridmetrics", "no band is empty in every cell", length(nms_all_na) == 0, if(length(nms_all_na)) paste(nms_all_na, collapse = ";") else "none", "none")

  #5. check per-cell return counts against R
  #  - the raster is compared cell-by-cell in row/column order against a
  #    reference grid built on the tile's own coordinates, so the count
  #    check still works even if the raster's coordinates are wrong
  nms_rcount = grep("^r[0-9]count$", names(r_gm), value = TRUE)
  v_gm_total = rowSums(terra::values(r_gm[[nms_rcount]]), na.rm = TRUE)
  r_ref_10 = fn_reference_grid(10)
  v_ref_10 = terra::values(terra::rasterize(v_pts, r_ref_10, fun = "length"))[ , 1]
  v_ref_10[is.na(v_ref_10)] = 0
  same_shape = terra::nrow(r_gm) == terra::nrow(r_ref_10) & terra::ncol(r_gm) == terra::ncol(r_ref_10)
  fn_check(10, "gridmetrics", "grid has the expected rows x columns for the tile", same_shape, paste(dim(r_gm)[1:2], collapse = "x"), paste(dim(r_ref_10)[1:2], collapse = "x"))
  if(same_shape) fn_check(10, "gridmetrics", "per-cell return counts match R", all(v_gm_total == v_ref_10), sum(v_gm_total != v_ref_10), "0 mismatched cells")

  #6. apply the height ceiling to elev_max
  v_gm_max = terra::values(r_gm[["elev_max"]])[ , 1]
  n_gm_too_tall = sum(v_gm_max > height_limit, na.rm = TRUE)
  fn_check(10, "gridmetrics", "no elev_max cell exceeds the tallest point above ground", n_gm_too_tall == 0, paste0(n_gm_too_tall, " cells; max ", round(max(v_gm_max, na.rm = TRUE), 2)), paste0("0 cells above ", round(height_limit, 2), " ft"))

  #7. check the CSV table agrees with the raster, cell by cell
  #  - the table is sorted by row then column to match raster cell order
  if(!is.na(path_gm_csv) && file.exists(path_gm_csv)) {
    dt_gm = data.table::fread(path_gm_csv)
    data.table::setorder(dt_gm, Row, Col)
    fn_check(10, "gridmetrics", "CSV table has one row per raster cell", nrow(dt_gm) == terra::ncell(r_gm), nrow(dt_gm), terra::ncell(r_gm))
    if(nrow(dt_gm) == terra::ncell(r_gm)) {
      v_gm_mean = terra::values(r_gm[["elev_mean"]])[ , 1]
      same_mean = all(abs(dt_gm$elev_mean - v_gm_mean) < 1e-3 | (is.na(dt_gm$elev_mean) & is.na(v_gm_mean)))
      fn_check(10, "gridmetrics", "CSV elev_mean matches the raster band", same_mean, sum(!same_mean), "all cells equal")
      }
    }

  #8. map elev_max, outlining cells above the height ceiling
  print(fn_plot_raster_map(r_gm, "elev_max", "gridmetrics elev_max (red = above the height ceiling)", highlight_above = height_limit))
  }
```

# Step 11: Summarize the tile and two plots with `cloudmetrics`

`cloudmetrics` writes one row of summary metrics for the whole cloud, or
one row per polygon with `/shape`. Both are checked against R's counts and
the height ceiling.

```{r}
#| label: step-11-cloudmetrics

#1. whole-tile metrics with height strata
path_cm_tile = file.path(dir_run, "cloudmetrics_tile.csv")
fn_run_tool(11, "cloudmetrics", c(path_las, paste0("/ground:", path_ground_dem), paste0("/output:", path_cm_tile), "/minht:2", "/strata:0.5,2,5,10,20"), "cloudmetrics tile")
if(fn_check_file(11, "cloudmetrics", path_cm_tile, "tile metrics CSV")) {
  dt_cm_tile = data.table::fread(path_cm_tile)
  print(dt_cm_tile[ , 1:12])
  fn_check(11, "cloudmetrics", "tile: total points matches the tile", dt_cm_tile$TotalPoints == n_points, dt_cm_tile$TotalPoints, n_points)
  fn_check(11, "cloudmetrics", "tile: elev_max within the tallest point above ground", dt_cm_tile$elev_max <= height_limit, round(dt_cm_tile$elev_max, 2), paste0("<= ", round(height_limit, 2), " ft"))
  n_canopy_expected = sum(height_above_fusion_ground > 2, na.rm = TRUE)
  fn_check(11, "cloudmetrics", "tile: canopy points (> 2 ft) match R (within 1%)", abs(dt_cm_tile$CanopyPoints - n_canopy_expected) <= max(3, 0.01 * n_canopy_expected), dt_cm_tile$CanopyPoints, n_canopy_expected)
  }

#2. per-plot metrics labelled by plot_id
path_cm_plots = file.path(dir_run, "cloudmetrics_plots.csv")
fn_run_tool(11, "cloudmetrics", c(path_las, paste0("/ground:", path_ground_dem), paste0("/output:", path_cm_plots), paste0("/shape:", path_plots_shp), "/field:plot_id"), "cloudmetrics /shape")
if(fn_check_file(11, "cloudmetrics", path_cm_plots, "per-plot metrics CSV")) {
  dt_cm_plots = data.table::fread(path_cm_plots)
  print(dt_cm_plots[ , 1:8])
  fn_check(11, "cloudmetrics", "/shape: one row per plot, labelled from /field", setequal(dt_cm_plots$Label, names(n_in_plot)), paste(dt_cm_plots$Label, collapse = ","), paste(names(n_in_plot), collapse = ","))
  for(plot_i in intersect(dt_cm_plots$Label, names(n_in_plot))) {
    n_cm_i = dt_cm_plots[Label == plot_i, TotalPoints]
    fn_check(11, "cloudmetrics", paste0("/shape: ", plot_i, " point count within 0.5% of R"), abs(n_cm_i - n_in_plot[[plot_i]]) <= max(3, 0.005 * n_in_plot[[plot_i]]), n_cm_i, n_in_plot[[plot_i]])
    }
  }
```

# Step 12: Compute slope and aspect with `topometrics`

`topometrics` derives slope and aspect from a DEM. It is run twice: on the
FUSION2 ground DEM from Step 5, and on the external USGS DTM of the same
ground, which has a coordinate system attached. The reference slope and
aspect come from
`terra::terrain()`. Edge cells have no full neighborhood, so they are left
out of the comparison.

```{r}
#| label: step-12-topometrics

#1. define a helper that runs topometrics on one DEM and compares it with
#   terra::terrain() on interior cells
fn_check_topometrics = function(path_dem, path_out, run_label){
  fn_run_tool(12, "topometrics", c(path_dem, paste0("/output:", path_out)), run_label)
  if(!fn_check_file(12, "topometrics", path_out, paste0(run_label, " raster"))) return(invisible(NULL))
  r_dem = terra::rast(path_dem)
  r_topo = terra::rast(path_out)
  print(fn_raster_overview(r_topo))

  #  - keep the input DEM's grid and coordinate system
  same_grid = all(abs(as.vector(terra::ext(r_topo)) - as.vector(terra::ext(r_dem))) < 1e-6) & all(dim(r_topo)[1:2] == dim(r_dem)[1:2])
  fn_check(12, "topometrics", paste0(run_label, ": output on the input DEM's grid"), same_grid, round(as.vector(terra::ext(r_topo)), 2), round(as.vector(terra::ext(r_dem)), 2))
  fn_check(12, "topometrics", paste0(run_label, ": keeps the input's coordinate system"), nzchar(terra::crs(r_topo)) == nzchar(terra::crs(r_dem)), if(nzchar(terra::crs(r_topo))) "present" else "none", if(nzchar(terra::crs(r_dem))) "present" else "none", severity = "soft")

  #  - interior cells only
  interior = terra::rast(r_dem)
  terra::values(interior) = 1
  interior[c(1, terra::nrow(r_dem)), ] = NA
  interior[ , c(1, terra::ncol(r_dem))] = NA
  keep = !is.na(terra::values(interior)[ , 1])

  #  - slope within 0.5 degrees
  slope_ref = terra::values(terra::terrain(r_dem, v = "slope", unit = "degrees"))[ , 1]
  slope_tool = terra::values(r_topo[["slope_degrees"]])[ , 1]
  diff_slope = abs(slope_tool - slope_ref)[keep]
  fn_check(12, "topometrics", paste0(run_label, ": interior cells have values"), sum(is.na(slope_tool[keep])) == 0, sum(is.na(slope_tool[keep])), 0)
  fn_check(12, "topometrics", paste0(run_label, ": median |slope - terra| under 0.5 degrees"), stats::median(diff_slope, na.rm = TRUE) < 0.5, round(stats::median(diff_slope, na.rm = TRUE), 3), "< 0.5")

  #  - aspect compared around the circle, only where the slope is steep
  #    enough (> 2 degrees) for aspect to be stable
  #  - terra reports a compass bearing (0 = north, clockwise); the tool may
  #    instead use the math convention (0 = east, counterclockwise), which
  #    is converted to a bearing as 90 - aspect. The values are correct if
  #    either reading matches; the soft check records which one it uses
  aspect_ref = terra::values(terra::terrain(r_dem, v = "aspect", unit = "degrees"))[ , 1]
  aspect_tool = terra::values(r_topo[["aspect_degrees"]])[ , 1]
  fn_angle_diff = function(a, b){
    d = abs((a - b) %% 360)
    pmin(d, 360 - d)
    }
  keep_aspect = keep & slope_ref > 2
  diff_compass = stats::median(fn_angle_diff(aspect_tool, aspect_ref)[keep_aspect], na.rm = TRUE)
  diff_math = stats::median(fn_angle_diff((90 - aspect_tool) %% 360, aspect_ref)[keep_aspect], na.rm = TRUE)
  fn_check(12, "topometrics", paste0(run_label, ": aspect matches terra (compass or math convention) within 5 degrees"), min(diff_compass, diff_math) < 5, paste0("compass ", round(diff_compass, 2), "; math ", round(diff_math, 2)), "< 5 under one convention")
  fn_check(12, "topometrics", paste0(run_label, ": aspect reported as a compass bearing (0 = north, clockwise)"), diff_compass < 5, if(diff_compass < 5) "compass bearing" else if(diff_math < 5) "math convention (0 = east, counterclockwise)" else "neither", "compass bearing", severity = "soft")

  #  - plot tool slope (x) against the terra reference (y)
  dt_cmp = data.table::data.table(topometrics = slope_tool[keep], terra_reference = slope_ref[keep])
  print(
    ggplot2::ggplot(dt_cmp, ggplot2::aes(x = topometrics, y = terra_reference)) +
      ggplot2::geom_abline(slope = 1, intercept = 0, colour = "grey50") +
      ggplot2::geom_point(alpha = 0.3, size = 0.6) +
      ggplot2::coord_equal() +
      ggplot2::labs(title = paste0("Slope: ", run_label), x = "topometrics slope (degrees)", y = "terra::terrain slope (degrees)") +
      ggplot2::theme_minimal(base_size = 11)
    )
  invisible(r_topo)
  }

#2. run on the FUSION2 ground DEM and the external USGS DTM
fn_check_topometrics(path_ground_dem, file.path(dir_run, "topometrics_ground.tif"), "topometrics ground DEM")
fn_check_topometrics(path_dtm_external, file.path(dir_run, "topometrics_external_dtm.tif"), "topometrics external DTM")
```

# Step 13: Compute surface roughness with `gridsurfacestats`

`gridsurfacestats` derives surface area ratio and roughness from a surface
raster. Given a second surface with `/reference`, it also computes the
cut/fill volume between them. It runs on the canopy height model with the
smoothed model as the reference, so `volume_diff` should equal the
per-cell height difference times cell area.

```{r}
#| label: step-13-gridsurfacestats

#1. run gridsurfacestats with a reference surface
path_surf = file.path(dir_run, "gridsurfacestats.tif")
fn_run_tool(13, "gridsurfacestats", c(path_chm, paste0("/output:", path_surf), paste0("/reference:", path_chm_smooth)), "gridsurfacestats")

if(fn_check_file(13, "gridsurfacestats", path_surf, "surface stats raster")) {
  #2. overview and placement
  r_surf = terra::rast(path_surf)
  print(fn_raster_overview(r_surf))
  fn_check_raster_extent(13, "gridsurfacestats", r_surf, "surface stats raster")
  fn_check_raster_crs(13, "gridsurfacestats", r_surf, "surface stats raster")

  #3. check value ranges that hold by definition
  #  - a surface's 3D area can never be less than its flat area, so the
  #    ratio is always >= 1
  v_sar = terra::values(r_surf[["surface_area_ratio"]])[ , 1]
  v_rough = terra::values(r_surf[["roughness"]])[ , 1]
  fn_check(13, "gridsurfacestats", "surface_area_ratio >= 1 everywhere", all(v_sar >= 1 - 1e-9, na.rm = TRUE), round(min(v_sar, na.rm = TRUE), 4), ">= 1")
  fn_check(13, "gridsurfacestats", "roughness >= 0 everywhere", all(v_rough >= 0, na.rm = TRUE), round(min(v_rough, na.rm = TRUE), 4), ">= 0")

  #4. check volume_diff against the per-cell height difference
  #  - which input is subtracted from which isn't documented, so either
  #    sign is accepted
  #  - the hard check accepts the difference with or without the cell-area
  #    factor; the soft check records whether the band is a true volume
  #    (height difference x cell area), as its name says
  if("volume_diff" %in% names(r_surf)) {
    v_vol = terra::values(r_surf[["volume_diff"]])[ , 1]
    v_height_diff = terra::values(r_chm)[ , 1] - terra::values(terra::rast(path_chm_smooth))[ , 1]
    cell_area_surf = prod(terra::res(r_chm))
    fn_matches = function(expected){
      all(abs(v_vol - expected) < 1e-3, na.rm = TRUE) | all(abs(v_vol + expected) < 1e-3, na.rm = TRUE)
      }
    is_volume = fn_matches(v_height_diff * cell_area_surf)
    is_depth = fn_matches(v_height_diff)
    fn_check(13, "gridsurfacestats", "volume_diff tracks the surface - reference difference", is_volume | is_depth, if(is_volume) "height difference x cell area" else if(is_depth) "height difference only" else "neither", "one of the two")
    fn_check(13, "gridsurfacestats", "volume_diff includes cell area (a true volume)", is_volume, if(is_volume) "volume" else if(is_depth) "height difference only (no cell area)" else "neither", "height difference x cell area", severity = "soft")
    } else {
    fn_check(13, "gridsurfacestats", "volume_diff band written with /reference", FALSE, "absent", "present")
    }
  }
```

# Step 14: Find tree tops with `canopymaxima`

`canopymaxima` finds local height peaks on the canopy height model. Each
reported tree top should sit inside the tile, stand at least `/minht`
tall, match the CHM's height at that location, and stay under the height
ceiling.

```{r}
#| label: step-14-canopymaxima

#1. run canopymaxima on the plain CHM
path_tops = file.path(dir_run, "tree_tops.csv")
fn_run_tool(14, "canopymaxima", c(path_chm, paste0("/output:", path_tops), "/minht:2"), "canopymaxima")

if(fn_check_file(14, "canopymaxima", path_tops, "tree tops CSV")) {
  dt_tops = data.table::fread(path_tops)
  print(head(dt_tops))
  fn_check(14, "canopymaxima", "at least one tree top found", nrow(dt_tops) > 0, nrow(dt_tops), "> 0")

  #2. location, minimum height, and height ceiling
  inside = dt_tops$X >= ext_tile[["xmin"]] - cellsize_chm & dt_tops$X <= ext_tile[["xmax"]] + cellsize_chm & dt_tops$Y >= ext_tile[["ymin"]] - cellsize_chm & dt_tops$Y <= ext_tile[["ymax"]] + cellsize_chm
  fn_check(14, "canopymaxima", "every tree top inside the tile", all(inside), sum(!inside), "0 outside")
  fn_check(14, "canopymaxima", "every tree top at least /minht (2 ft)", all(dt_tops$Height >= 2), round(min(dt_tops$Height), 2), ">= 2")
  n_tops_too_tall = sum(dt_tops$Height > height_limit)
  fn_check(14, "canopymaxima", "no tree top exceeds the tallest point above ground", n_tops_too_tall == 0, paste0(n_tops_too_tall, " tops; max ", round(max(dt_tops$Height), 2)), paste0("0 above ", round(height_limit, 2), " ft"))

  #3. check the CSV writes coordinates at full precision
  #  - a value like 2.05097e+06 keeps only 6 significant digits, which
  #    moves a tree top by up to 5 ft on this tile's coordinates
  lines_tops = readLines(path_tops)[-1]
  n_sci = sum(grepl("[0-9]e[+-][0-9]", lines_tops))
  fn_check(14, "canopymaxima", "X/Y written without scientific-notation rounding", n_sci == 0, paste0(n_sci, " rows, e.g. X = ", strsplit(lines_tops[1], ",")[[1]][2]), "0 rows")

  #4. each top's height should equal the CHM cell it sits in
  #  - a mismatch usually means the reported X/Y lost precision, so the
  #    top points at the wrong cell
  chm_at_top = terra::extract(r_chm, cbind(dt_tops$X, dt_tops$Y))[ , 1]
  n_mismatch = sum(abs(chm_at_top - dt_tops$Height) > 0.01 | is.na(chm_at_top))
  fn_check(14, "canopymaxima", "each top's height equals the CHM at its X/Y", n_mismatch == 0, n_mismatch, "0 mismatched tops")
  }
```

# Step 15: Segment tree crowns with `treeseg`

`treeseg` grows a crown region around each peak on the CHM. It writes a
segment raster and a per-crown table. The table's pixel counts and maximum
heights should match what R reads from the segment raster and the CHM.

```{r}
#| label: step-15-treeseg

#1. run treeseg on the plain CHM
path_crowns_tif = file.path(dir_run, "crown_segments.tif")
path_crowns_csv = file.path(dir_run, "crown_summary.csv")
fn_run_tool(15, "treeseg", c(path_chm, paste0("/output-raster:", path_crowns_tif), paste0("/output-table:", path_crowns_csv), "/minht:2"), "treeseg")

if(fn_check_file(15, "treeseg", path_crowns_tif, "segment raster") & fn_check_file(15, "treeseg", path_crowns_csv, "crown table")) {
  r_crowns = terra::rast(path_crowns_tif)
  dt_crowns = data.table::fread(path_crowns_csv)
  print(head(dt_crowns))
  fn_check_raster_extent(15, "treeseg", r_crowns, "segment raster")
  fn_check_raster_crs(15, "treeseg", r_crowns, "segment raster")

  #2. rebuild the per-crown pixel count and max height from the rasters
  dt_crown_ref = data.table::data.table(TreeID = terra::values(r_crowns)[ , 1], chm = terra::values(r_chm)[ , 1])
  dt_crown_ref = dt_crown_ref[!is.na(TreeID), .(PixelCount_ref = .N, MaxHeight_ref = max(chm, na.rm = TRUE)), by = TreeID]
  dt_crown_cmp = merge(dt_crowns, dt_crown_ref, by = "TreeID", all = TRUE)

  #3. compare the table with the rebuilt values
  fn_check(15, "treeseg", "same crowns in the table and the raster", nrow(dt_crown_cmp) == nrow(dt_crowns) & nrow(dt_crown_cmp) == nrow(dt_crown_ref), nrow(dt_crowns), nrow(dt_crown_ref))
  fn_check(15, "treeseg", "PixelCount matches the segment raster", all(dt_crown_cmp$PixelCount == dt_crown_cmp$PixelCount_ref, na.rm = FALSE), sum(dt_crown_cmp$PixelCount != dt_crown_cmp$PixelCount_ref, na.rm = TRUE), "0 mismatched crowns")
  fn_check(15, "treeseg", "MaxHeight matches the CHM within each crown", all(abs(dt_crown_cmp$MaxHeight - dt_crown_cmp$MaxHeight_ref) < 0.01), sum(abs(dt_crown_cmp$MaxHeight - dt_crown_cmp$MaxHeight_ref) >= 0.01, na.rm = TRUE), "0 mismatched crowns")
  n_crowns_too_tall = sum(dt_crowns$MaxHeight > height_limit)
  fn_check(15, "treeseg", "no crown exceeds the tallest point above ground", n_crowns_too_tall == 0, paste0(n_crowns_too_tall, " crowns; max ", round(max(dt_crowns$MaxHeight), 2)), paste0("0 above ", round(height_limit, 2), " ft"))

  #4. crown area should be pixel count x cell area
  cell_area_chm = prod(terra::res(r_chm))
  col_area = grep("CrownArea", names(dt_crowns), value = TRUE)[1]
  vec_area = dt_crowns[[col_area]]
  fn_check(15, "treeseg", "crown area = pixel count x cell area", all(abs(vec_area - dt_crowns$PixelCount * cell_area_chm) < 1), sum(abs(vec_area - dt_crowns$PixelCount * cell_area_chm) >= 1), "0 mismatched crowns")
  fn_check(15, "treeseg", "crown area column's unit label matches the tile's units (feet)", !any(grepl("_m2$", names(dt_crowns))), grep("Area", names(dt_crowns), value = TRUE), "no m2 label on a feet-based tile", severity = "soft")

  #5. map the crowns
  print(fn_plot_raster_map(r_crowns, 1, "treeseg crown segments (IDs)"))
  }
```

# Step 16: Chain tools per tile with `pipeline`

`pipeline` splits the input into tiles with a buffer, runs a chain of
tools on each tile, and mosaics the results. It needs an explicit
`/extent` (script 003 Step 2 explains why). The chain here is ground DEM,
then canopy model, then tree tops, on four 100 x 100 ft tiles. The
mosaicked canopy model should match the single-run model from Step 7
wherever both are plausible.

```{r}
#| label: step-16-pipeline

#1. run a three-stage chain over four tiles
dir_pl = file.path(dir_run, "pipeline")
arg_extent_tile = paste0("/extent:", paste(ext_tile[c("xmin", "ymin", "xmax", "ymax")], collapse = ","))
fn_run_tool(16, "pipeline", c(
    "/pipeline:groundfilter,canopymodel,canopymaxima"
  , paste0("/input:", path_las)
  , paste0("/output:", dir_pl)
  , arg_extent_tile
  , "/tilesize:100,100"
  , "/buffer:20"
  , "/cellsize:3"
  , "/minht:2"
  , "/merge"
  ), "pipeline")

#2. check the run state table: every tile x stage should have finished
path_pl_state = file.path(dir_pl, "_processing", "pipeline_state.csv")
if(fn_check_file(16, "pipeline", path_pl_state, "run state table")) {
  dt_pl_state = data.table::fread(path_pl_state, colClasses = "character")
  print(dt_pl_state[ , .N, by = .(stage, status)])
  fn_check(16, "pipeline", "no tile x stage failed", !any(grepl("fail", dt_pl_state$status, ignore.case = TRUE)), paste(unique(dt_pl_state$status), collapse = ","), "no failed status")
  fn_check(16, "pipeline", "four tiles processed", data.table::uniqueN(dt_pl_state$tile) == 4, data.table::uniqueN(dt_pl_state$tile), 4)
  }

#3. compare the mosaicked CHM with Step 7's single-run CHM
#  - the merged GeoTIFF is preferred; the VRT mosaic is used if /merge
#    produced nothing
paths_pl_chm = list.files(dir_pl, pattern = "canopymodel.*\\.(tif|vrt)$", full.names = TRUE)
paths_pl_chm = paths_pl_chm[order(!grepl("\\.tif$", paths_pl_chm))]
if(fn_check(16, "pipeline", "mosaicked canopy model written", length(paths_pl_chm) > 0, if(length(paths_pl_chm)) basename(paths_pl_chm[1]) else "none", "a canopymodel .tif or .vrt")) {
  r_pl_chm = terra::rast(paths_pl_chm[1])
  fn_check_raster_extent(16, "pipeline", r_pl_chm, "mosaicked CHM")
  r_pl_chm_aligned = try(terra::resample(r_pl_chm, r_chm, method = "near"), silent = TRUE)
  if(!inherits(r_pl_chm_aligned, "try-error")) {
    v_pl = terra::values(r_pl_chm_aligned)[ , 1]
    keep_pl = !is.na(v_pl) & !is.na(v_chm) & v_chm <= height_limit & v_pl <= height_limit
    diff_pl = abs(v_pl[keep_pl] - v_chm[keep_pl])
    fn_check(16, "pipeline", "mosaicked CHM matches single-run CHM (median diff < 0.5 ft)", length(diff_pl) > 0 && stats::median(diff_pl) < 0.5, if(length(diff_pl)) round(stats::median(diff_pl), 3) else "no overlapping cells", "< 0.5 ft")
    n_pl_too_tall = sum(terra::values(r_pl_chm)[ , 1] > height_limit, na.rm = TRUE)
    fn_check(16, "pipeline", "no mosaicked CHM cell exceeds the tallest point above ground", n_pl_too_tall == 0, n_pl_too_tall, 0)
    } else {
    fn_check(16, "pipeline", "mosaicked CHM can be aligned with the single-run CHM", FALSE, "resample failed", "aligned")
    }
  }

#4. check tree tops were concatenated across tiles without duplicates at
#   tile seams
#  - a duplicate is the same X, Y, and height reported twice; matching on
#    X and Y alone would also flag distinct trees whose coordinates were
#    rounded to the same value (see Step 14)
paths_pl_tops = list.files(dir_pl, pattern = "canopymaxima.*\\.csv$", full.names = TRUE)
if(fn_check(16, "pipeline", "combined tree tops table written", length(paths_pl_tops) > 0, length(paths_pl_tops), ">= 1")) {
  dt_pl_tops = data.table::fread(paths_pl_tops[1])
  n_dup_tops = sum(duplicated(dt_pl_tops[ , .(X, Y, Height)]))
  fn_check(16, "pipeline", "no duplicate tree tops across tile seams", n_dup_tops == 0, n_dup_tops, 0)
  }
```

# Step 17: Gather every check into one table

This step collects every check from Steps 1-16, prints the failures, and
saves the full log as a versioned CSV. The chart shows each tool's pass,
warn, and fail counts at a glance. Each failed `hard` check is a candidate
for a unit test in `FUSION2/tests/cpp` that reproduces it before the
fix.

```{r}
#| label: step-17-summary

#1. assemble the log and label each result
dt_checks = data.table::rbindlist(env_checks$rows)
dt_checks[ , result := data.table::fifelse(passed, "pass", data.table::fifelse(severity == "hard", "fail", "warn"))]

#2. print a per-tool tally and every non-passing check
dt_tally = data.table::dcast(dt_checks[ , .N, by = .(tool, result)], tool ~ result, value.var = "N", fill = 0)
print(dt_tally)
print(dt_checks[result != "pass", .(step, tool, check, observed, expected, result)])

#3. save the log, stamped with the tool build it tested
dt_checks[ , build_modified := dt_build$modified[match(tool, dt_build$tool)]]
path_check_log = RForInvt::file_version(
    file.path(dir_tables, "all_tools_check_log.csv")
  , increment = TRUE
  , note = "per-tool output checks from 005_example_all_tools.qmd"
  )
data.table::fwrite(dt_checks, path_check_log)
cat("Check log written to:", path_check_log, "\n")

#4. save every command this run executed, in run order, with its exit
#   status and the run folder it wrote into
dt_commands = data.table::rbindlist(env_checks$commands)
dt_commands[ , run_folder := basename(dir_run)]
path_command_log = RForInvt::file_version(
    file.path(dir_tables, "all_tools_commands.csv")
  , increment = TRUE
  , note = "every tool command run by 005_example_all_tools.qmd"
  )
data.table::fwrite(dt_commands, path_command_log)
cat("Command log written to:", path_command_log, "\n")

#5. chart pass/warn/fail counts per tool, in the order the tools ran
dt_plot = dt_checks[ , .N, by = .(tool, result)]
dt_plot[ , tool := factor(tool, levels = rev(unique(dt_checks$tool)))]
dt_plot[ , result := factor(result, levels = c("fail", "warn", "pass"))]
ggplot2::ggplot(dt_plot, ggplot2::aes(x = N, y = tool, fill = result)) +
  ggplot2::geom_col(width = 0.7) +
  ggplot2::scale_fill_manual(values = c(fail = "#C0392B", warn = "#E0A030", pass = "#5B8C5A")) +
  ggplot2::labs(title = "Output checks per tool", x = "Number of checks", y = NULL, fill = NULL) +
  ggplot2::theme_minimal(base_size = 11)
```

# Session Info

```{r}
#| label: session-info

sessionInfo()
```
