---
title: "Prepare a Forested Example Lidar Tile"
author: "Jacob Strunk"
date: today
format: html
execute:
  echo: true
  warning: false
  message: false
---

<!--
Pipeline script -- downloads one public USGS 3DEP lidar tile, finds its
most heavily forested 200 x 200 ft window, and clips that window into
data/raw/lidar/ as the example tile for 005_example_all_tools.qmd. Follows
the pipeline template (execution-ordered "# Step N" headings).
-->

# Setup

## Reset working directory to the project root

This script lives in `R/analysis/`, but Quarto runs chunks from the
`.qmd`'s own folder, so project-relative paths need the working directory
moved to the project root. The setting takes effect from the next chunk.

```{r}
#| label: root-dir

knitr::opts_knit$set(root.dir = "../..")
```

## Load libraries and set paths

```{r}
#| label: setup

#1. load packages
library(data.table)
library(terra)
library(lidR)
library(ggplot2)

#2. set the source tile and output locations
#  - the full tile is large (tens to hundreds of MB) and regenerable, so it
#    is cached under data/raw/lidar_full_tiles/, which git ignores; only
#    the small clip written to data/raw/lidar/ is committed
#  - the tile's download URL comes from the USGS National Map products
#    service (tnmaccess.nationalmap.gov/api/v1/products, dataset "Lidar
#    Point Cloud (LPC)") queried at the forest site's coordinates
#  - a tile that was obtained some other way is used by setting url_tile
#    to NA and placing the file in data/raw/lidar_full_tiles/ under
#    name_tile; Step 1 then skips the download
url_tile = NA
name_tile = "TC_1372.laz"
dir_full_tiles = file.path("data", "raw", "lidar_full_tiles")
dir_lidar_raw = file.path("data", "raw", "lidar")
dir.create(dir_full_tiles, recursive = TRUE, showWarnings = FALSE)
if(!is.na(url_tile)) name_tile = basename(url_tile)
path_full_tile = file.path(dir_full_tiles, name_tile)
path_forest_clip = file.path(dir_lidar_raw, sub("\\.laz$", "_forest200ft.laz", name_tile))

#3. set the window-selection parameters, in feet
#  - Step 2 converts them into the tile's own linear unit (feet or metres)
window_size_ft   = 200   # clip width and height
cellsize_scan_ft = 10    # grid cell used to score candidate windows
height_cover_ft  = 6     # a first return above this height counts as canopy
```

# Step 1: Download the full tile

The tile is fetched once and reused on later runs. A download that stops
early leaves a truncated file, so the file's size is checked against the
server's reported size before it is used.

```{r}
#| label: step-1-download

#1. download the tile if it has a URL and is not already cached
if(!is.na(url_tile) && !file.exists(path_full_tile)) {
  utils::download.file(url_tile, path_full_tile, mode = "wb", quiet = TRUE)
  }
if(!file.exists(path_full_tile)) stop("Full tile not found at ", path_full_tile)

#2. interrogate: a downloaded file must match the server's size
if(!is.na(url_tile)) {
  size_local = file.info(path_full_tile)$size
  headers_remote = curlGetHeaders(url_tile)
  line_length = grep("^content-length:", headers_remote, ignore.case = TRUE, value = TRUE)
  size_remote = as.numeric(gsub("[^0-9]", "", line_length[length(line_length)]))
  cat("Local size:", size_local, "bytes; server size:", size_remote, "bytes\n")
  if(size_local != size_remote) stop("Cached tile is incomplete -- delete ", path_full_tile, " and rerun.")
  }
```

# Step 2: Grid canopy cover and height across the tile

Heights above ground are computed against a triangulation of the tile's
own ground points. On a 10 ft grid, each cell gets its canopy cover (the
share of first returns above 6 ft) and its 95th-percentile height.

```{r}
#| label: step-2-grid-canopy

#1. read the tile and normalize heights
las_full = lidR::readLAS(path_full_tile, filter = "-drop_withheld")
las_full_norm = lidR::normalize_height(las_full, algorithm = lidR::tin())

#2. convert the window parameters into the tile's linear unit
#  - US survey feet and international feet differ by 2 ppm, which does
#    not matter at these sizes, so both count as feet
unit_tile = sf::st_crs(las_full)$units_gdal
unit_is_feet = grepl("foot|feet|ft", unit_tile, ignore.case = TRUE)
unit_per_ft = if(unit_is_feet) 1 else 0.3048
cat("Tile linear unit:", unit_tile, "\n")
if(!unit_is_feet & !grepl("metre|meter", unit_tile, ignore.case = TRUE)) stop("Unrecognized tile unit: ", unit_tile)
window_size   = window_size_ft * unit_per_ft
cellsize_scan = cellsize_scan_ft * unit_per_ft
height_cover  = height_cover_ft * unit_per_ft

#3. compute per-cell canopy cover and 95th-percentile height
fn_cell_canopy = function(z, rn){
  z_first = z[rn == 1]
  list(
      cover = if(length(z_first) > 0) mean(z_first > height_cover) else NA_real_
    , p95   = as.numeric(stats::quantile(z, 0.95))
    )
  }
r_canopy = lidR::pixel_metrics(las_full_norm, ~fn_cell_canopy(Z, ReturnNumber), res = cellsize_scan)

#4. interrogate: the grid must cover the tile and hold plausible heights
print(r_canopy)
v_p95 = terra::values(r_canopy[["p95"]])[ , 1]
cat("95th-percentile height across cells (tile units):\n")
print(stats::quantile(v_p95, c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE))
if(max(v_p95, na.rm = TRUE) > 400 * unit_per_ft) stop("Heights above 400 ft -- normalization failed.")
```

# Step 3: Score every 200 ft window and pick the most forested one

Every window of 21 x 21 cells (210 ft) is scored by its mean canopy
cover and mean 95th-percentile height. The scoring window has an odd
width so it has a true centre cell (`terra::focal` requires this). The
200 ft clip is then centred on that cell, so it sits wholly inside the
scored area. Windows that run off the tile or contain empty cells are
skipped. The chosen window is the tallest one among those with at least
70% canopy cover.

```{r}
#| label: step-3-pick-window

#1. average cover and height over every 21 x 21 cell window
n_cells_window = round(window_size / cellsize_scan) + 1
kernel = matrix(1, n_cells_window, n_cells_window)
r_cover_mean = terra::focal(r_canopy[["cover"]], w = kernel, fun = "mean", na.rm = FALSE)
r_p95_mean   = terra::focal(r_canopy[["p95"]], w = kernel, fun = "mean", na.rm = FALSE)

#2. collect the scored windows and centre a 200 ft clip on each
dt_windows = data.table::data.table(
    cell  = seq_len(terra::ncell(r_canopy))
  , cover = terra::values(r_cover_mean)[ , 1]
  , p95   = terra::values(r_p95_mean)[ , 1]
  )
dt_windows = dt_windows[!is.na(cover) & !is.na(p95)]
xy_centre = terra::xyFromCell(r_canopy, dt_windows$cell)
dt_windows[ , xmin := xy_centre[ , 1] - window_size / 2]
dt_windows[ , ymin := xy_centre[ , 2] - window_size / 2]
dt_windows[ , xmax := xmin + window_size]
dt_windows[ , ymax := ymin + window_size]

#3. keep windows fully inside the tile, then take the tallest dense one
ext_full = terra::ext(r_canopy)
dt_windows = dt_windows[xmin >= ext_full$xmin & xmax <= ext_full$xmax & ymin >= ext_full$ymin & ymax <= ext_full$ymax]
dt_dense = dt_windows[cover >= 0.70]
if(nrow(dt_dense) == 0) stop("No 200 ft window on this tile has 70% canopy cover -- choose another tile.")
window_best = dt_dense[which.max(p95)]
print(window_best)

#4. interrogate: map the tile's canopy height with the chosen window
dt_map = data.table::as.data.table(terra::as.data.frame(r_canopy[["p95"]], xy = TRUE))
p_window = ggplot2::ggplot(dt_map, ggplot2::aes(x = x, y = y, fill = p95)) +
  ggplot2::geom_raster() +
  ggplot2::annotate("rect", xmin = window_best$xmin, xmax = window_best$xmax, ymin = window_best$ymin, ymax = window_best$ymax, fill = NA, colour = "red", linewidth = 1) +
  ggplot2::scale_fill_viridis_c(name = "95th pct\nheight (ft)") +
  ggplot2::coord_equal() +
  ggplot2::labs(title = "Full tile canopy height; red box = chosen example window", x = NULL, y = NULL) +
  ggplot2::theme_minimal()
print(p_window)
```

# Step 4: Clip the window and write the example tile

The window is clipped from the full tile, withheld points included, and
written as compressed LAZ in the tile's own LAS version and point format,
so the example exercises the same encoding USGS distributes. The clip's
own ground points must still be present, since every height-based check
in 005 depends on them.

```{r}
#| label: step-4-clip

#1. clip from the unfiltered tile so withheld points stay in the example
las_source = lidR::readLAS(path_full_tile)
las_clip = lidR::clip_rectangle(las_source, window_best$xmin, window_best$ymin, window_best$xmax, window_best$ymax)
lidR::writeLAS(las_clip, path_forest_clip)

#2. interrogate the written clip
las_check = lidR::readLAS(path_forest_clip)
las_check_norm = lidR::normalize_height(lidR::filter_poi(las_check, !Withheld_flag), algorithm = lidR::tin())
dt_clip_summary = data.table::data.table(
    n_points     = lidR::npoints(las_check)
  , n_ground     = sum(las_check$Classification == 2)
  , n_withheld   = sum(las_check$Withheld_flag)
  , cover        = mean(las_check_norm$Z[las_check_norm$ReturnNumber == 1] > height_cover)
  , height_p95   = as.numeric(stats::quantile(las_check_norm$Z, 0.95))
  , height_max   = max(las_check_norm$Z)
  , las_version  = paste0(las_check@header$`Version Major`, ".", las_check@header$`Version Minor`)
  , point_format = las_check@header$`Point Data Format ID`
  , file_mb      = round(file.info(path_forest_clip)$size / 1e6, 2)
  )
print(dt_clip_summary)
if(dt_clip_summary$n_ground < 100) stop("Clip has fewer than 100 ground points -- heights cannot be normalized reliably.")
if(is.na(sf::st_crs(las_check))) stop("Clip lost its coordinate system.")
if(dt_clip_summary$file_mb > 5) warning("Clip is larger than 5 MB -- consider a smaller window before committing it.")
```

# Step 5: Fetch a matching USGS bare-earth DTM for the clip

005 runs `topometrics` a second time on an external DTM, to check that the
tool keeps an input raster's coordinate system. That DTM should cover the
same ground as the example tile, so its slope and aspect can be compared
with those from FUSION2's own ground surface. The USGS 3DEP elevation
service returns a bare-earth clip for any extent, reprojected to the
tile's coordinate system.

The service reports elevations in metres. They are converted to the
tile's own vertical unit before saving, since a DTM whose elevations and
coordinates are in different units gives wrong slopes.

```{r}
#| label: step-5-fetch-dtm

#1. set the request: the clip's extent plus a 50 ft pad on every side, on
#   a 3 ft grid, in the clip's horizontal coordinate system
#  - terra reports the horizontal part of the tile's compound (horizontal
#    plus vertical) coordinate system as an EPSG code the service accepts
pad_dtm = 50 * unit_per_ft
cellsize_dtm = 3 * unit_per_ft
ext_dtm = c(
    xmin = floor(las_check@header$`Min X` - pad_dtm)
  , ymin = floor(las_check@header$`Min Y` - pad_dtm)
  , xmax = ceiling(las_check@header$`Max X` + pad_dtm)
  , ymax = ceiling(las_check@header$`Max Y` + pad_dtm)
  )
crs_desc = terra::crs(terra::rast(ext = terra::ext(0, 1, 0, 1), crs = sf::st_crs(las_check)$wkt), describe = TRUE)
epsg_dtm = crs_desc$code
if(is.na(epsg_dtm) || !nzchar(epsg_dtm)) stop("Could not find an EPSG code for the clip's horizontal coordinate system.")
n_cols_dtm = round((ext_dtm[["xmax"]] - ext_dtm[["xmin"]]) / cellsize_dtm)
n_rows_dtm = round((ext_dtm[["ymax"]] - ext_dtm[["ymin"]]) / cellsize_dtm)
url_dtm = paste0(
    "https://elevation.nationalmap.gov/arcgis/rest/services/3DEPElevation/ImageServer/exportImage"
  , "?bbox=", paste(ext_dtm[c("xmin", "ymin", "xmax", "ymax")], collapse = ",")
  , "&bboxSR=", epsg_dtm, "&imageSR=", epsg_dtm
  , "&size=", n_cols_dtm, ",", n_rows_dtm
  , "&format=tiff&pixelType=F32&interpolation=RSP_BilinearInterpolation&f=image"
  )

#2. download the clip, retrying a few times
#  - the service occasionally times out on a first request
dir_dtm_site = file.path("data", "raw", "lidar_dtm_examples", "site_tc_1372_forest")
dir.create(dir_dtm_site, recursive = TRUE, showWarnings = FALSE)
path_dtm_download = file.path(tempdir(), "usgs_3dep_dtm_metres.tif")
r_dtm_m = NULL
for(attempt_i in 1:4) {
  result_i = try(utils::download.file(url_dtm, path_dtm_download, mode = "wb", quiet = TRUE), silent = TRUE)
  if(!inherits(result_i, "try-error")) r_dtm_m = try(terra::rast(path_dtm_download), silent = TRUE)
  if(!is.null(r_dtm_m) && !inherits(r_dtm_m, "try-error")) break
  r_dtm_m = NULL
  Sys.sleep(10)
  }
if(is.null(r_dtm_m)) stop("USGS 3DEP elevation service did not return a readable raster after 4 attempts.")

#3. convert elevations from metres to the tile's vertical unit and save
#  - US survey feet: 1 m = 3937/1200 ft
metres_to_unit = if(unit_is_feet) 3937 / 1200 else 1
r_dtm = r_dtm_m * metres_to_unit
names(r_dtm) = "elevation"
path_dtm = file.path(dir_dtm_site, "usgs_3dep_dtm.tif")
terra::writeRaster(r_dtm, path_dtm, overwrite = TRUE, gdal = c("COMPRESS=LZW"))

#4. interrogate: the DTM must sit on the requested extent in the clip's
#   coordinate system, have no gaps, and agree with the clip's own ground
#   points to within a few feet
r_dtm_check = terra::rast(path_dtm)
print(r_dtm_check)
if(terra::crs(r_dtm_check, describe = TRUE)$code != epsg_dtm) stop("Saved DTM is not in EPSG:", epsg_dtm)
if(any(abs(as.vector(terra::ext(r_dtm_check)) - ext_dtm[c("xmin", "xmax", "ymin", "ymax")]) > cellsize_dtm)) stop("Saved DTM does not cover the requested extent.")
if(any(is.na(terra::values(r_dtm_check)))) stop("Saved DTM has empty cells.")
las_ground_check = lidR::filter_poi(las_check, Classification == 2L & !Withheld_flag)
z_dtm_at_ground = terra::extract(r_dtm_check, cbind(las_ground_check$X, las_ground_check$Y), method = "bilinear")[ , 1]
diff_dtm = las_ground_check$Z - z_dtm_at_ground
cat("Clip ground points minus USGS DTM (tile units): median", round(stats::median(diff_dtm), 2), "; 95% of points within", round(stats::quantile(abs(diff_dtm), 0.95), 2), "\n")
if(abs(stats::median(diff_dtm)) > 3 * unit_per_ft) stop("USGS DTM and the clip's ground points differ by more than 3 ft at the median -- check units and vertical datum.")

#5. plot the clip's ground points (y) against the DTM beneath them (x)
dt_dtm_cmp = data.table::data.table(usgs_dtm = z_dtm_at_ground, ground_points = las_ground_check$Z)
range_dtm_cmp = range(c(dt_dtm_cmp$usgs_dtm, dt_dtm_cmp$ground_points))
ggplot2::ggplot(dt_dtm_cmp, ggplot2::aes(x = usgs_dtm, y = ground_points)) +
  ggplot2::geom_abline(slope = 1, intercept = 0, colour = "grey50") +
  ggplot2::geom_point(alpha = 0.3, size = 0.6) +
  ggplot2::coord_fixed(ratio = 1, xlim = range_dtm_cmp, ylim = range_dtm_cmp) +
  ggplot2::labs(title = "Clip ground points vs. USGS 3DEP DTM", x = "USGS DTM elevation (tile units)", y = "ground point elevation (tile units)") +
  ggplot2::theme_minimal(base_size = 11)
```
