---
title: "Export the All-Tools Run's Figures, Commands, and Checks to the FUSION2 Manual"
author: "Jacob Strunk"
date: today
format: html
execute:
  echo: true
  warning: false
  message: false
---

<!--
Pipeline script -- takes one finished 005 run (its run folder, check log,
and command log) and writes everything the FUSION2 user manual's worked
examples need into the sibling FUSION2 repository's docs/examples/ folder:

  docs/examples/figures/    one PNG per figure in 005_example_all_tools_REPORT.qmd
  docs/examples/tables/     the check log, the command log, and run details as CSV
  docs/examples/fragments/  small Markdown pieces the manual's chapters
                            include: each tool's commands, checks, and figures

The figures are drawn by 005_example_all_tools_REPORT.qmd itself, rendered
to Markdown so Quarto saves each figure as a PNG named by its chunk label.
No plotting code is copied here. PNG is used instead of this project's
usual TIFF because neither the manual's HTML pages nor its PDF can show
TIFF images.

The files written into FUSION2 are versioned by FUSION2's git history, not
by RForInvt::file_version(): each export replaces the previous one, and
the manual always shows the latest run.

Run 005 first, then this script, then rebuild the manual in FUSION2
(build.ps1). The FUSION2 repository is expected next to this one
(../FUSION2); set the FUSION2_REPO_DIR environment variable to use a
different location.
-->

# Setup

## Reset working directory to the project root

This script lives in `R/analysis/`, but Quarto runs chunks from the
`.qmd`'s own folder, so project-relative paths need the working directory
moved to the project root. The setting takes effect from the next chunk.

```{r}
#| label: root-dir

knitr::opts_knit$set(root.dir = "../..")
```

## Load libraries and set default options

Loads `data.table` for the logs and `RForInvt` to find the newest
versioned logs.

```{r}
#| label: setup

#1. load packages
library(data.table)
library(RForInvt)

#2. set project directories
dir_tables    = file.path("output", "tables")
dir_runs_root = file.path("data", "derived", "all_tools_example")
dir_scratch   = file.path("data", "derived", "docs_export")
path_report   = file.path("R", "analysis", "005_example_all_tools_REPORT.qmd")

#3. locate the FUSION2 repository and its manual's examples folder
dir_fusion2 = Sys.getenv("FUSION2_REPO_DIR", unset = file.path("..", "FUSION2"))
if(!file.exists(file.path(dir_fusion2, "build.ps1"))) stop("FUSION2 repository not found at ", dir_fusion2, "; set FUSION2_REPO_DIR")
dir_docs_examples = file.path(dir_fusion2, "docs", "examples")
dir_out_figures   = file.path(dir_docs_examples, "figures")
dir_out_tables    = file.path(dir_docs_examples, "tables")
dir_out_fragments = file.path(dir_docs_examples, "fragments")
for(dir_i in c(dir_out_figures, dir_out_tables, dir_out_fragments)) dir.create(dir_i, recursive = TRUE, showWarnings = FALSE)

#4. set the tool order the manual follows (the order 005 runs them in)
nms_tools = c(
    "catalog", "filterdata", "thindata", "groundfilter", "clipdata"
  , "canopymodel", "returndensity", "densitymetrics", "gridmetrics"
  , "cloudmetrics", "topometrics", "gridsurfacestats", "canopymaxima"
  , "treeseg", "pipeline"
  )
```

## Define helper functions used throughout this script

These helpers turn log rows into Markdown the manual can include.

```{r}
#| label: helper-functions

#1. define a helper that shortens a long value for a table cell, and
#   escapes the "|" character Markdown tables use as a column divider
fn_cell = function(x, max_chars = 45){
  x_in = trimws(as.character(x))
  x_in[is.na(x_in)] = ""
  too_long = nchar(x_in) > max_chars
  x_in[too_long] = paste0(substr(x_in[too_long], 1, max_chars - 3), "...")
  gsub("|", "\\|", x_in, fixed = TRUE)
  }

#2. define a helper that writes a data.table as a Markdown pipe table,
#   with a Quarto caption and label so the table is numbered
fn_pipe_table = function(dt, caption, label){
  header = paste0("| ", paste(names(dt), collapse = " | "), " |")
  divider = paste0("|", paste(rep("---", ncol(dt)), collapse = "|"), "|")
  rows = apply(dt, 1, function(r) paste0("| ", paste(r, collapse = " | "), " |"))
  c(header, divider, rows, "", paste0(": ", caption, " {#", label, "}"), "")
  }

#3. define a helper that lays one command out for the Windows Command
#   Prompt: the tool and its input on the first line, then one option per
#   line, joined with the "^" line-continuation character
#  - splitting on " /" keeps an option's value with its option; the
#    commands never contain " /" inside a value (paths use forward slashes
#    with no leading space)
fn_command_block = function(command){
  parts = strsplit(command, " /", fixed = TRUE)[[1]]
  if(length(parts) == 1) return(parts)
  c(paste0(parts[1], " ^"), paste0("    /", parts[-c(1, length(parts))], " ^"), paste0("    /", parts[length(parts)]))
  }
```

# Step 1: Pick the 005 run to export

Picks the newest 005 command log and the check log from the same run,
then reads the run folder named in the command log. Both logs are written
at the end of one 005 run, so they share a run folder.

```{r}
#| label: step-1-pick-run

#1. read the newest command log and check log
path_command_log = RForInvt::file_version(file.path(dir_tables, "all_tools_commands.csv"))
path_check_log   = RForInvt::file_version(file.path(dir_tables, "all_tools_check_log.csv"))
dt_commands = data.table::fread(path_command_log)
dt_checks   = data.table::fread(path_check_log)

#2. find the run folder the commands wrote into
run_folder = unique(dt_commands$run_folder)
if(length(run_folder) != 1) stop("Command log names more than one run folder: ", paste(run_folder, collapse = ", "))
dir_run = file.path(dir_runs_root, "run", run_folder)
if(!dir.exists(dir_run)) stop("Run folder not found: ", dir_run)

#3. confirm the two logs come from the same run
#  - both are stamped when 005's last step runs, so their timestamps
#    (the trailing _YYYYMMDDHHMMSS) match
stamp_commands = sub(".*_([0-9]{14})\\.csv$", "\\1", basename(path_command_log))
stamp_checks   = sub(".*_([0-9]{14})\\.csv$", "\\1", basename(path_check_log))
if(stamp_commands != stamp_checks) stop("Command log (", stamp_commands, ") and check log (", stamp_checks, ") come from different 005 runs")

#4. every command must have exited cleanly and every hard check passed
#   before the run is used as the manual's example
if(any(dt_commands$exit_status != 0)) stop("Some commands in this run did not exit cleanly")
n_failed = dt_checks[result == "fail", .N]
if(n_failed > 0) stop(n_failed, " hard checks failed in this run; fix them before exporting it to the manual")

cat("Exporting run:", dir_run, "\n")
cat("Commands:", nrow(dt_commands), " Checks:", nrow(dt_checks), "\n")
```

# Step 2: Render the report's figures to PNG

Renders `005_example_all_tools_REPORT.qmd` to Markdown for the chosen run.
Quarto saves each figure as a PNG named after its chunk label
(`fig-catalog-density-1.png`), which is all this step keeps from the
render.

```{r}
#| label: step-2-render-figures

#1. clear the previous render
dir_render = file.path(dir_scratch, "render")
unlink(dir_render, recursive = TRUE)
dir.create(dir_render, recursive = TRUE, showWarnings = FALSE)

#2. render the report to Markdown for this run
#  - Quarto resolves --output-dir relative to the report's own folder
args_render = c(
    "render", path_report
  , "--to", "gfm"
  , "--output-dir", file.path("..", "..", dir_render)
  , "-M", "fig-dpi:150"
  , "-P", paste0("dir_run:", dir_run)
  , "-P", paste0("path_check_log:", path_check_log)
  )
render_status = system2("quarto", args = args_render)
if(render_status != 0) stop("Rendering the report failed with status ", render_status)

#3. find the rendered figures
dir_render_figures = file.path(dir_render, "005_example_all_tools_REPORT_files", "figure-commonmark")
paths_png = list.files(dir_render_figures, pattern = "\\.png$", full.names = TRUE)
cat(length(paths_png), "figures rendered\n")
```

# Step 3: Match each figure to its caption and tool

Reads every figure label and caption from the report's source, so the
manual shows the same captions. The tool is the word after `fig-` in the
label.

```{r}
#| label: step-3-captions

#1. read label and caption pairs from the report source
lines_report = readLines(path_report, warn = FALSE)
idx_label = grep("^#\\| label: fig-", lines_report)
dt_figures = data.table::data.table(
    label   = sub("^#\\| label: ", "", lines_report[idx_label])
  , caption = sub("^#\\| fig-cap: \"(.*)\"$", "\\1", lines_report[idx_label + 1])
  )

#2. name each figure's tool and PNG
dt_figures[ , tool := sub("^fig-([a-z]+).*$", "\\1", label)]
dt_figures[ , path_png := file.path(dir_render_figures, paste0(label, "-1.png"))]

#3. every figure in the report must have a PNG and belong to a known tool
missing_png = dt_figures[!file.exists(path_png), label]
if(length(missing_png) > 0) stop("No PNG rendered for: ", paste(missing_png, collapse = ", "))
unknown_tool = dt_figures[!tool %in% nms_tools, label]
if(length(unknown_tool) > 0) stop("Figure label does not start with a tool name: ", paste(unknown_tool, collapse = ", "))
if(any(!grepl("^fig-", dt_figures$label)) || any(grepl("fig-cap", dt_figures$caption))) stop("A figure label is not followed by its caption in the report source")
print(dt_figures[ , .(label, tool)])
```

# Step 4: Copy figures and logs into the FUSION2 manual

Replaces the manual's figures and tables with this run's. Old files are
cleared first so a figure dropped from the report doesn't linger in the
manual; FUSION2's git history keeps every earlier export.

```{r}
#| label: step-4-copy

#1. replace the figures
unlink(list.files(dir_out_figures, pattern = "\\.png$", full.names = TRUE))
path_png_out = file.path(dir_out_figures, paste0(dt_figures$label, ".png"))
copied = file.copy(dt_figures$path_png, path_png_out, overwrite = TRUE)
if(!all(copied)) stop("Could not copy: ", paste(dt_figures$label[!copied], collapse = ", "))

#2. copy the two logs
data.table::fwrite(dt_checks, file.path(dir_out_tables, "all_tools_check_log.csv"))
data.table::fwrite(dt_commands, file.path(dir_out_tables, "all_tools_commands.csv"))

#3. record which run and which FUSION2 build the manual's examples show
dt_run_info = data.table::data.table(
    item  = c("run folder", "command log", "check log", "tools tested (modified)", "commands", "checks", "checks passed", "checks warned")
  , value = c(
        run_folder
      , basename(path_command_log)
      , basename(path_check_log)
      , max(dt_checks$build_modified, na.rm = TRUE)
      , nrow(dt_commands)
      , nrow(dt_checks)
      , dt_checks[result == "pass", .N]
      , dt_checks[result == "warn", .N]
      )
  )
data.table::fwrite(dt_run_info, file.path(dir_out_tables, "run_info.csv"))
print(dt_run_info)
```

# Step 5: Write the Markdown pieces the manual includes

Writes three pieces per tool (its commands, its checks, and its figures)
plus three for the appendix (run details, a tally by tool, and every
check). The manual's chapters live in `docs/tools/`, one folder away
from `docs/examples/`, so figure paths are written relative to there.

```{r}
#| label: step-5-fragments

#1. clear the previous pieces
unlink(list.files(dir_out_fragments, pattern = "\\.md$", full.names = TRUE))

#2. write each tool's commands, checks, and figures
for(tool_i in nms_tools){
  #  - commands: one Command Prompt block, a REM line naming each run
  dt_cmd_i = dt_commands[tool == tool_i]
  lines_cmd_i = c("```bat")
  for(j in seq_len(nrow(dt_cmd_i))){
    if(j > 1) lines_cmd_i = c(lines_cmd_i, "")
    lines_cmd_i = c(lines_cmd_i, paste("REM", dt_cmd_i$run_label[j]), fn_command_block(dt_cmd_i$command[j]))
    }
  lines_cmd_i = c(lines_cmd_i, "```", "")
  writeLines(lines_cmd_i, file.path(dir_out_fragments, paste0(tool_i, "_commands.md")))

  #  - checks: one numbered table
  dt_chk_i = dt_checks[tool == tool_i, .(
      Check    = fn_cell(check, 70)
    , Result   = result
    , Observed = fn_cell(observed)
    , Expected = fn_cell(expected)
    )]
  lines_chk_i = fn_pipe_table(
      dt_chk_i
    , caption = paste0("Automated checks on the example run of `", tool_i, "`.")
    , label = paste0("tbl-checks-", tool_i)
    )
  writeLines(lines_chk_i, file.path(dir_out_fragments, paste0(tool_i, "_checks.md")))

  #  - figures: one numbered figure per report figure, same caption
  dt_fig_i = dt_figures[tool == tool_i]
  lines_fig_i = character(0)
  for(j in seq_len(nrow(dt_fig_i))){
    lines_fig_i = c(lines_fig_i, paste0("![", dt_fig_i$caption[j], "](../examples/figures/", dt_fig_i$label[j], ".png){#", dt_fig_i$label[j], "}"), "")
    }
  writeLines(lines_fig_i, file.path(dir_out_fragments, paste0(tool_i, "_figures.md")))
  }

#3. write the appendix pieces
#  - run details
writeLines(fn_pipe_table(dt_run_info[ , .(Item = item, Value = fn_cell(value, 60))], "The example run the manual's worked examples come from.", "tbl-run-info"), file.path(dir_out_fragments, "run_info.md"))

#  - tally by tool, in run order ("reference" is 005's own setup step)
dt_tally = data.table::dcast(dt_checks[ , .N, by = .(tool, result)], tool ~ result, value.var = "N", fill = 0)
for(col_i in setdiff(c("pass", "warn", "fail"), names(dt_tally))) dt_tally[ , (col_i) := 0L]
dt_tally = dt_tally[match(unique(dt_checks$tool), tool), .(Tool = tool, Checks = pass + warn + fail, Pass = pass, Warn = warn, Fail = fail)]
writeLines(fn_pipe_table(dt_tally, "Automated check results by tool on the example run.", "tbl-checks-by-tool"), file.path(dir_out_fragments, "checks_by_tool.md"))

#  - every check
dt_all = dt_checks[ , .(Tool = tool, Check = fn_cell(check, 60), Severity = severity, Result = result, Observed = fn_cell(observed, 30), Expected = fn_cell(expected, 30))]
writeLines(fn_pipe_table(dt_all, "Every automated check on the example run, in the order it ran.", "tbl-all-checks"), file.path(dir_out_fragments, "all_checks.md"))
```

# Step 6: Check the export

Confirms that every tool got a commands piece with at least one command,
a checks piece with at least one check, and that every figure the
fragments point to exists.

```{r}
#| label: step-6-check

#1. every tool has commands and checks
n_cmd_by_tool = dt_commands[ , .N, by = tool]
n_chk_by_tool = dt_checks[ , .N, by = tool]
missing_cmd = setdiff(nms_tools, n_cmd_by_tool$tool)
missing_chk = setdiff(nms_tools, n_chk_by_tool$tool)
if(length(missing_cmd) > 0) stop("No commands recorded for: ", paste(missing_cmd, collapse = ", "))
if(length(missing_chk) > 0) stop("No checks recorded for: ", paste(missing_chk, collapse = ", "))

#2. every figure a fragment points to exists in the manual's figures folder
paths_fragments = list.files(dir_out_fragments, pattern = "_figures\\.md$", full.names = TRUE)
lines_fragments = unlist(lapply(paths_fragments, readLines))
refs = unlist(regmatches(lines_fragments, gregexpr("figures/[a-z-]+\\.png", lines_fragments)))
missing_fig = refs[!file.exists(file.path(dir_docs_examples, refs))]
if(length(missing_fig) > 0) stop("Fragments point to missing figures: ", paste(missing_fig, collapse = ", "))

#3. summarize what was written
data.table::data.table(
    folder = c("figures", "tables", "fragments")
  , files  = c(length(list.files(dir_out_figures)), length(list.files(dir_out_tables)), length(list.files(dir_out_fragments)))
  )
```

# Session Info

```{r}
#| label: session-info

sessionInfo()
```
